---
title: "03_preprocessing"
author: "Roger Calvas"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      fig.path = "../figures/03_PreProcessing/")
```

#remove any potential contaminants and evaluate the accuracy of our sequencing run. Then, we eill write our final data objects.
1. Load ni data that we´ve generated: ´asv_table´, ´tax_table´, and metadata (which includes ´metadata.csv´ and ´track_reads_reads´). We created these files in ´analysis/02_AssignASVs.Rmd´.
2. Remove any ASVs that are cloroplasts.
3. Remove ASVs that are mitochondria.
4. Evaluate any ASVs from the negative control. Then, remove negative controls.
5. Evaluate the mock community (positive control). Accuracy of sequencing.
6. Check for reverse complements.
7. Check the sequencing depth of our samples. Remove obvious samples with too few reads.

## Input

1. Metadata : ´metadata.csv´ and ´track_read_counts.Rdata´
2. ASV table : ´data/01_DADA2/ASV_table.csv´
3. Taxonomy Table: ´data/01_DADA2/ASV_taxonomy.tsv´

##output

1. A pre-processed phyloseq object (S4 object): ´raw_preprocessed_physeq.RData´.

#Set Environment

## Load Packages
```{r load-packages}
#install.packages("BiocManager")
#BiocManager::install("Biostrings")
pacman::p_load(devtools, phyloseq, Biostrings, tidyverse, install = FALSE)
```

## Load Data

### 1. Metadata
-What to copy? 

1. # file.copy('/workdir/in_class_data/metadata.csv',         '/workdir/rc969/SalinityGradient_16s/data/metadata.csv')


2. #file.copy('/workdir/in_class_data/SalinityGradient_16s/track_read_counts.RDat                a', '/workdir/rc969/SalinityGradient_16s/data/01_DADA2/track_read_c                   ounts.RData')




```{r load-metadata}
metadata_df <- read_csv("data/metadata/SraRunTable.csv", col_names = TRUE) %>%
  #fix column Name
  dplyr::rename("sample_names" = "Run") %>%
  # Add sample names as column
  mutate(names = sample_names)
  
#Inspect
  View(metadata_df)
  head(metadata_df)
  
# include data2 output
  load("data/01_DADA2/track_read_counts.RData")

#take a look
  head(track_counts_df)

  # Fix sample names in track_reads
  track_counts_df$sample_names <- sapply(strsplit(track_counts_df$sample_names, "_"), '[', 1)
  
  #Intuition check
  
#Whats the different 
  setdiff(track_counts_df$sample_names, metadata_df$sample_names)
  # do a filtering join with left_join
  metadata_final_df <-
    metadata_df %>%
    left_join(., track_counts_df, by = "sample_names") %>%
    #sample names to the rownames to merge into phyloseq
    column_to_rownames(var = "sample_names")
#check
  dim(metadata_final_df)
```

## 2. ASV table
copy data over to = 'workdir/rc969/SalinityGradient_16s/data/01_DADA02/' 
  
```{r load-asv-table}
asv_df <-
  read.delim(file = "data/01_DADA2/ASV_table.tsv",  sep = "\t",
           #add column names and row names
           header = TRUE, row.names = 1) 
 
#Inspect
asv_df[1:3, 1:3]

#fix column names
## Remove the x: denote at the beginning "^"
colnames(asv_df) <- sub(pattern = "^X", replacement = "", colnames(asv_df))
##Replace the . with a - : "\\." Since . is a special character in regex
gsub(pattern = "\\.", replacement = "-" , colnames(asv_df))

#Final inspection
head(colnames(asv_df))
asv_df[1:3, 1:3]
```
 
 ## 3. Taxonomy Table
 
```{r}
tax_df <-
  read.delim("data/01_DADA2/ASV_taxonomy.tsv", sep = "\t",
             header = TRUE, row.names = 1)
#inspect
dim(tax_df)
dim(asv_df)

#add check 
stopifnot(rownames(asv_df) == rownames(tax_df))
```

# Handof to phyloseq

```{r phyloseq-handoff}
raw_physeq <-
  phyloseq::phyloseq(otu_table(asv_df, taxa_are_rows = TRUE),
         sample_data(metadata_final_df),
         tax_table(as.matrix(tax_df)))

raw_physeq

#save
save(raw_physeq, file = "data/03_PreProcessing/raw_physeq.RData")
```

#Clean up the data!

Remove chloroplasts and mitochondria
```{r rm-chloro-mitos}
noChloros_physeq <-
  raw_physeq %>%
    subset_taxa(Order != "Chloroplast" | is.na(Order))

# How many ASVs were chloroplasts?
numChloros_ASVs <- ntaxa(raw_physeq) - ntaxa(noChloros_physeq)
```

  There were 'r numChloros_ASVs' ASVs that were chloroplasts.
  
What about mitochondria?
```{r rm-mitos}
 noChlorosMitos_physeq <-
  noChloros_physeq %>%
    subset_taxa(Family != "Mitochondria" | is.na(Family))

# how many ASVs were mitochondrias?
numMitos_ASVs <-  ntaxa(noChloros_physeq) - ntaxa(noChlorosMitos_physeq)
```

  There were 'numMitos_ASVs' ASVs that were mitochondrias.
#Evaluate and remove the control samples

```{r neg-controls}
# create a vector of samples that were negative controls 
control_samples 

# Make a new phyloseq object of only the control samples 
control_physeq <- 
  noChlorosMitos_physeq %>%
  subset_samples(., names %in% control_samples) %>%
  # force remove ASVs with 0 counts 
  prune_taxa(taxa_sums(.) > 0, .)
  
# Let's take a look
control_physeq
```