---
title: "04_APhylogenetic_Tree_Construction"
author: "Roger Calvas"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      fig.path = "../../../figures/04_Phylogenetic_Tree/")
```

#Goals of this file
##The goal is to make a phylogenetic tree!

1.Load in preprocessed phyloseq object
2.Create ASV fasta file from the phyloseq object
3.Align the 16S sequences from fasta file with MAFT
4.Create a tree with FastTree2


#Set Environment
##Set seed
```{r set-seed}
# Set the seed for reproducibility
set.seed(39809473)

# Let's make a parameter to set the number of threads 
n_threads = 20
```

#Timing of the script
```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
```
## Load Packages

```{r load-packages}
# phytools, ggtree, RColorBrewer
pacman::p_load(phytools, ggtree, RColorBrewer, install = FALSE)

# install phytools
#remotes::install_github("liamrevell/phytools")

# install ggtree
#devtools::install_github("GuangchuangYu/ggtree")

# Load physeq 
load("data/03_PreProcessing/raw_preprocessed_physeq.RData")

raw_preprocessed_physeq <-
  noControlsMock_ChlorosMitos_physeq

raw_preprocessed_physeq
```

#Write ASV fasta-file
```{r create-preprocessed-fasta}
# Pull out ASV sequences and ASV names
asv_seq_df <- 
  raw_preprocessed_physeq@tax_table %>% 
  data.frame() %>% 
  dplyr::select(ASV, ASVseqs)

#View(asv_seq_df)

# Add the ">" to make fasta header
asv_seq_df$ASV <- paste0(">", asv_seq_df$ASV)

#View(asv_seq_df)

# Create fasta object
asv_seq_fasta <- c(rbind(asv_seq_df$ASV, asv_seq_df$ASVseqs))
head(asv_seq_fasta)

# Write to a file 
write(asv_seq_fasta, 
      file = "data/04_Phylogenetic_Tree/preprocessed_ASVs.fasta")
```

# Allign the 16s sequences from fasta file with MAFFT
- 'engine.opts = '-l'': gives us our original terminal base environment

Test shell comands in R markdown file
- ctrl + alt + enter
```{r run-maft, engine = 'bash', engine.opts = '-l'}
# Write bash code to run MAFFT
# First provide the path to MAFT
export PATH=/programs/mafft/bin:$PATH
# change directories to provide the fasta file we made above 
cd data/04_Phylogenetic_Tree/
pwd


# Set a seed  - using same seed as before for consistency 
RANDOM=39809473

# Run MAFFT
# For now, use default options, note the version 
# MAFFT automatically knows that it's a nucleotide alignment 
/programs/mafft/bin/mafft --auto preprocessed_ASVs.fasta > MAFFT_aligned_ASVs.fasta
```

#FastTree2
```{r run-fasttree, engine  ='bash', engine.opts = '-l'}
# Provide export path to fasttree 
export PATH=/programs/FastTree-2.1.11:$PATH

# cd into the alignment file folder 
cd data/04_Phylogenetic_Tree/
pwd

# Run Fasttree to generate phylogenetic tree 
# parameters: 
    # -nt = indicates it's a nucleotide alignment
    # -gtr = generalized time reversible substitution model 
    # -fastest speed up the model, reduce memory usage (recommended for datasets that have >50,000)
    # -log = output a log file 
    # input alignment file 
    # specify the output tree file 
FastTree -nt -gtr -fastest -log FastTree.log MAFFT_aligned_ASVs.fasta > ASVs_unrooted.tree
```


