---
title: "03_preprocessing"
author: "Roger Calvas"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      fig.path = "../../figures/03_PreProcessing/")
```

# Set Environment

## Load Packages
```{r load-packages}
#install.packages("BiocManager")
#BiocManager::install("Biostrings")
pacman::p_load(devtools, phyloseq, Biostrings, tidyverse, patchwork, install = FALSE)
```

## Set seed
```{r set-seed}
# Set the seed for reproducibility
set.seed(9473)

# Let's make a parameter to set the number of threads 
n_threads = 20
```

## Timing of the script
```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
```

# Objetivos del archivo:

1. Remove any ASVs that are **cloroplasts**
2. Remove ASVs that are **mitochondria**
3. Check for **reverse complements**
4. Check los **sequencing depth** y decidir acerca
del acerca del minimo para rarefaccion
5. Finalizar el archivo physeq `preprocessed_physeq`

## Input
1. **Raw phyloseq object**: `data/02_AssignASVs/raw_phyloseq.RData`

## Output
1. A pre-processed phyloseq object (S4 object): `data/03_PreProcessing/preprocessed_phyloseq.RData`
# Load Data

```{r load-physeq}
# Load physeq
load("data/its/02_Assign_ASVs/raw_physeq.RData")

# Ver el formato
raw_physeq
```

# Clean up the data!

## Remover todo lo no_fungi
```{r rm-!fungi}
# Mantener estrictamente solo a los ASVs asignados al reino Fungi
fungi_physeq <-
  raw_physeq %>%
  subset_taxa(Kingdom == "k__Fungi")

# How many ASVs were non-fungi (e.g., host DNA, plants, protists)?
numNonFungi_ASVs <- ntaxa(raw_physeq) - ntaxa(fungi_physeq)

numNonFungi_ASVs

# Eliminar secuencias que no tienen asignacion a nivel de phylum 
clean_physeq <-
  fungi_physeq %>%
  subset_taxa(!is.na(Phylum))

# How many ambiguous ASVs were removed?
numAmbiguous_ASVs <- ntaxa(fungi_physeq) - ntaxa(clean_physeq)
numAmbiguous_ASVs
```

<span style="color: red;"> Pregunta 1: Que contaminantes buscaste y cuantos habian? El dataset tenia controles positivos o negativos? como se manejan los controles? </span> 
  
there is no controls nor mock communities so we can go ahead to Reverse complements

## Buscar complemento reverso
```{r reverse-complements}
# Pull out Taxa Table
tax_table <- 
  clean_physeq %>%
  tax_table() %>% 
  data.frame()

# Grab ASV Sequences and Names
asv_names <- tax_table$ASV
asv_seqs <- tax_table$ASVseqs

# Create empty comparison matrix
asv_rev_comp <- matrix(nrow=length(asv_seqs), ncol = length(asv_seqs)) 

# Fix the names in the rows and columns 
rownames(asv_rev_comp) <- asv_names
colnames(asv_rev_comp) <- asv_names

# Convert sequences to Biostrings
asv_dna <- DNAStringSet(asv_seqs) 

# Inspect 
head(asv_dna)

# Construct reverse complement
asv_rev_dna <- reverseComplement(asv_dna) 

# Now loop through every asv sequence to check 
for(i in 1:length(asv_seqs)){ # For each asv sequence...
  match_vec <- asv_dna[i] == asv_rev_dna # Compare it to the reverse complement of every other sequence...
  asv_rev_comp[,i] <- match_vec # Write that logical vector as a new column 
}

# Find how many TRUEs (matches) we have, divide by 2 because each pair occurs twice
cat("For", sum(asv_rev_comp) / 2,"ASVs, the reverse complement will need to be removed") 

# The current data objec
clean_physeq
```

# Asesorar la profundidad del resto de las muestras

```{r read-depth}
# El objeto actual
clean_physeq

# Make a metadata_df for joining
metadata_df <- 
  clean_physeq %>%
  sample_data() %>% 
  data.frame()

# What is the library size/sequencing depth for each sample? 
seqSums_df <- 
  clean_physeq %>%
  otu_table() %>%
  # Sum each sample column 
  colSums() %>%
  data.frame() %>%
  rownames_to_column(var = "sample_names") %>%
  left_join(., metadata_df, by = "sample_names")

# Rename second column 
colnames(seqSums_df)[2] <- "TotalSeqs"

# check
dim(seqSums_df)
head(seqSums_df)

# Show the depth of samples 
seqSums_df %>%
  dplyr::select(sample_names, TotalSeqs) %>%
  arrange(TotalSeqs) %>%
  head()

# plot it as a bar plot 
numSeq_bar_plot <-
seqSums_df %>%
  ggplot(aes(x=reorder(sample_names, TotalSeqs), y = TotalSeqs,
             fill = localidad)) + 
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_blank()) + 
  labs(y = "Read Depth", x = "Sample")

# histogram
numSeq_hist_plot <- 
  seqSums_df %>%
  ggplot(aes(x= TotalSeqs, fill = localidad)) + 
  geom_histogram(color = "black") + 
  labs(y = "# of Samples", x = "Read Depth") + 
  theme(legend.position = "bottom")

# Density plot 
numSeq_density_plot <-
seqSums_df %>%
  ggplot(aes(TotalSeqs, fill = localidad)) +
  geom_density(alpha = 0.5) + 
  labs(x = "Read Depth")


# Put it all together 
numSeq_bar_plot + numSeq_hist_plot + numSeq_density_plot + 
  plot_annotation(tag_levels = "A") 

# What's the min seq depth? 
min(sample_sums(clean_physeq))

# Final check of the sequencing depth for now
seqSums_df %>%
  dplyr::select(sample_names, TotalSeqs) %>%
  arrange(TotalSeqs)
```

<span style="color: red;">Pregunta 2: Como se ven las profundidades, crees que hay "suficientes" reads para el resto del analisis o hay alguna muestra que tenga demasiado poco? </span>

# Save proccesed phyloseq object
```{r save-object}
save(clean_physeq, file = "data/its/03_PreProcessing/clean_physeq.RData")
```

# Final info for Reproducibility 

## Check Render Time
```{r stop-time}
# Take the time now that we are at the end of the script
end_time <- Sys.time()
end_time 

# Echo the elapsed time
elapsed_time <- round((end_time - start_time), 3)
elapsed_time
```

## Session Information

```{r session-info}
# Ensure reproducibility with package version information
devtools::session_info()
```

