---
title: "03_preprocessing"
author: "Roger Calvas"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      fig.path = "../../figures/03_PreProcessing/")
```

#remove any potential contaminants and evaluate the accuracy of our sequencing run. Then, we eill write our final data objects.
1. Load in data that we´ve generated: ´asv_table´, ´tax_table´, and metadata (which includes ´metadata.csv´ and ´track_reads_reads´). We created these files in ´analysis/02_AssignASVs.Rmd´.
2. Remove any ASVs that are cloroplasts.
3. Remove ASVs that are mitochondria.
4. Evaluate any ASVs from the negative control. Then, remove negative controls.
5. Evaluate the mock community (positive control). Accuracy of sequencing.
6. Check for reverse complements.
7. Check the sequencing depth of our samples. Remove obvious samples with too few reads.

## Input

1. Metadata : ´metadata.csv´ and ´track_read_counts.Rdata´
2. ASV table : ´data/01_DADA2/ASV_table.csv´
3. Taxonomy Table: ´data/01_DADA2/ASV_taxonomy.tsv´

##output

1. A pre-processed phyloseq object (S4 object): ´raw_preprocessed_physeq.RData´.

#Set Environment

## Load Packages
```{r load-packages}
#install.packages("BiocManager")
#BiocManager::install("Biostrings")
pacman::p_load(devtools, phyloseq, Biostrings, tidyverse, patchwork, install = FALSE)
```

## Load Data

### 1. Metadata
-What to copy? 

1. # file.copy('/workdir/in_class_data/metadata.csv',         '/workdir/rc969/SalinityGradient_16s/data/metadata.csv')


2. #file.copy('/workdir/in_class_data/SalinityGradient_16s/track_read_counts.RDat                a', '/workdir/rc969/SalinityGradient_16s/data/01_DADA2/track_read_                    counts.RData')

#Set seed

```{r set-seed}
# Set the seed for reproducibility
set.seed(39809473)

# Let's make a parameter to set the number of threads 
n_threads = 20
```

#Timing of the script
```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
```


```{r load-metadata}
metadata_df <- read_csv("data/metadata/Metadatafall.csv", col_names = TRUE) %>%
  #fix column Name
  dplyr::rename("sample_names" = "Run") %>%
  # Add sample names as column
  mutate(names = sample_names)
  
#Inspect
#View(metadata_df)
  head(metadata_df)
  
# include data2 output
  load("data/01_DADA2/track_read_counts.RData")

#take a look
  head(track_counts_df)

  # Fix sample names in track_reads
  track_counts_df$sample_names <- sapply(strsplit(track_counts_df$sample_names, "_"), '[', 1)
  
  #Intuition check
  
#Whats the different 
  setdiff(track_counts_df$sample_names, metadata_df$sample_names)
  # do a filtering join with left_join
  metadata_final_df <-
    metadata_df %>%
    left_join(., track_counts_df, by = "sample_names") %>%
    #sample names to the rownames to merge into phyloseq
    column_to_rownames(var = "sample_names") 
#check
  dim(metadata_final_df)
```

## 2. ASV table
copy data over to = 'workdir/rc969/SalinityGradient_16s/data/01_DADA02/' 
  
```{r load-asv-table}
asv_df <-
  read.delim(file = "data/01_DADA2/ASV_table.tsv",  sep = "\t",
           #add column names and row names
           header = TRUE, row.names = 1) 
 
#Inspect
asv_df[1:3, 1:3]

#fix column names
## Remove the x: denote at the beginning "^"
colnames(asv_df) <- sub(pattern = "^X", replacement = "", colnames(asv_df))
##Replace the . with a - : "\\." Since . is a special character in regex
gsub(pattern = "\\.", replacement = "-" , colnames(asv_df))

#Final inspection
head(colnames(asv_df))
asv_df[1:3, 1:3]
```
 
 ## 3. Taxonomy Table
 
```{r}
tax_df <-
  read.delim("data/01_DADA2/ASV_taxonomy.tsv", sep = "\t",
             header = TRUE, row.names = 1)
#inspect
dim(tax_df)
dim(asv_df)

#add check 
stopifnot(rownames(asv_df) == rownames(tax_df))
```

# Handof to phyloseq

```{r phyloseq-handoff}
raw_physeq <-
  phyloseq::phyloseq(otu_table(asv_df, taxa_are_rows = TRUE),
         sample_data(metadata_final_df),
         tax_table(as.matrix(tax_df)))

raw_physeq

#save
save(raw_physeq, file = "data/03_PreProcessing/raw_physeq.RData")
```

#Clean up the data!

Remove chloroplasts and mitochondria
```{r rm-chloro-mitos}
noChloros_physeq <-
  raw_physeq %>%
    subset_taxa(Order != "Chloroplast" | is.na(Order))

# How many ASVs were chloroplasts?
numChloros_ASVs <- ntaxa(raw_physeq) - ntaxa(noChloros_physeq)
```

  There were 'r numChloros_ASVs' ASVs that were chloroplasts.
  
What about mitochondria?
```{r rm-mitos}
 noControlsMock_ChlorosMitos_physeq <-
  noChloros_physeq %>%
    subset_taxa(Family != "Mitochondria" | is.na(Family))

# how many ASVs were mitochondrias?
numMitos_ASVs <-  ntaxa(noChloros_physeq) - ntaxa(noControlsMock_ChlorosMitos_physeq)
```

  There were 'numMitos_ASVs' ASVs that were mitochondrias.
  
Interpretation #1:
  I got 10 chloroplasts and 1 mitochondria in my 1991 ASVs this is pretty reasonable since this is a gut sample and even surprising because there is almost no contamination from epitelial cells or plant matter.
  
#There is no controls nor mock communities so we can go ahead to Reverse complements

My study only had negative controls for the treatment with fiber and no negative control for sequencing or mock communities, since I subsetted my data as to only work with inulin (one of the treatment groups), there are no negative treatment controls.

```{r reverse-complements}
# Pull out Taxa Table
tax_table <- 
 noControlsMock_ChlorosMitos_physeq %>%
  tax_table() %>% 
  data.frame()

# Grab ASV Sequences and Names
asv_names <- tax_table$ASV
asv_seqs <- tax_table$ASVseqs

# Create empty comparison matrix
asv_rev_comp <- matrix(nrow=length(asv_seqs), ncol = length(asv_seqs)) 

# Fix the names in the rows and columns 
rownames(asv_rev_comp) <- asv_names
colnames(asv_rev_comp) <- asv_names

# Convert sequences to Biostrings
asv_dna <- DNAStringSet(asv_seqs) 

# Inspect 
head(asv_dna)

# Construct reverse complement
asv_rev_dna <- reverseComplement(asv_dna) 

# Now loop through every asv sequence to check 
for(i in 1:length(asv_seqs)){ # For each asv sequence...
  match_vec <- asv_dna[i] == asv_rev_dna # Compare it to the reverse complement of every other sequence...
  asv_rev_comp[,i] <- match_vec # Write that logical vector as a new column 
}

# Find how many TRUEs (matches) we have, divide by 2 because each pair occurs twice
cat("For", sum(asv_rev_comp) / 2,"ASVs, the reverse complement will need to be removed") 

# The current data objec
noControlsMock_ChlorosMitos_physeq

# What is the library size/sequencing depth for each sample? 
seqSums_df <- 
  noControlsMock_ChlorosMitos_physeq %>%
  otu_table() %>%
  # Sum each sample column 
  colSums() %>%
  data.frame() %>%
  rownames_to_column(var = "names") %>%
  left_join(., metadata_final_df, by = "names") 

# Rename second column 
colnames(seqSums_df)[2] <- "TotalSeqs"

# check
dim(seqSums_df)

head(seqSums_df)
# Show the depth of samples 
seqSums_df %>%
  dplyr::select(names, TotalSeqs) %>%
  arrange(TotalSeqs) %>%
  head()
# plot it as a bar plot 
numSeq_bar_plot <-
seqSums_df %>%
  ggplot(aes(x=reorder(names, TotalSeqs), y = TotalSeqs,
             fill = status)) + 
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_blank()) + 
  labs(y = "Read Depth", x = "Sample")

# histogram
numSeq_hist_plot <- 
  seqSums_df %>%
  ggplot(aes(x= TotalSeqs, fill = status)) + 
  geom_histogram(color = "black") + 
  labs(y = "# of Samples", x = "Read Depth") + 
  theme(legend.position = "bottom")

# Density plot 
numSeq_density_plot <-
seqSums_df %>%
  ggplot(aes(TotalSeqs, fill = status)) +
  geom_density(alpha = 0.5) + 
  labs(x = "Read Depth")


# Put it all together 
numSeq_bar_plot + numSeq_hist_plot + numSeq_density_plot + 
  plot_annotation(tag_levels = "A") 

# What's the min seq depth? 
min(sample_sums(noControlsMock_ChlorosMitos_physeq))

# Final check of the sequencing depth for now
# Here, we could remove more samples if we needed to
seqSums_df %>%
  dplyr::select(names, TotalSeqs) %>%
  arrange(TotalSeqs)
```

Since I am working with a gut microbiome dataset I need a good enough sequencing depth to look at the whole picture. My samples all have pretty good sequencing depth with the minimum being 7000 and more than half of them being over 16000 capping at 30000~, I think i don't need to remove any samples since non have bad sequencing depth and I think I have enough sequencing depth for further analysis mainly because more than half of them are around 20000 and looking at the literature this is a decent minimum in gut samples.

#Save proccesed phyloseq object
```{r save-object}
save(noControlsMock_ChlorosMitos_physeq, file = "data/03_PreProcessing/raw_preprocessed_physeq.RData")
```
#Final info for reproducibility
#Check render time
```{r render-time}
# Take the time now that we are at the end of the script
end_time <- Sys.time()
end_time 
# Echo the elapsed time
elapsed_time <- round((end_time - start_time), 3)
elapsed_time
# Ensure reproducibility with package version information
devtools::session_info()
```

