---
title: "04_BPhylogenetic_Tree"
author: "Roger Calvas"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      fig.path = "../../../figures/04_Phyylogenetic_Tree/")
```

#Goals
We want to investigate and root our tree

1.Load the FastTree unrooted tree
2. Add the tree to phyloseq object
3. Visualize and inspect tree with ggtree
4. Prune ASVs, if needed
5. Root out tree
6. Combine newly rooted tree with a phyloseq object 
7. Save 2 phyloseq objects:
  1. unrooted tree phyloseq object
  2. rooted tree phyloseq object
  
# Input
1. raw_preprocessed_physeq.RData
2. ASVs_unrooted.tree
3. MAFFT_aligned_ASVs.fasta

#output
1.phytree_preprocessed_physeq.RData

#output 
1. phytree_preprocessed_physeq.RData

#Before you start

##set my seed
```{r set-seed}
set.seed(39809473)
```
#Load packages
```{r load-packages}
pacman::p_load(tidyverse, phyloseq, ggtree, phytools, tidytree,
               install = FALSE)
```

#Load data files
```{r load-files}
# Preprocessed phyloseq object 
load("data/03_PreProcessing/raw_preprocessed_physeq.RData")
raw_preprocessed_physeq <-
  noControlsMock_ChlorosMitos_physeq
#raw_preprocessed_physeq
# Load in the tree! 
unrooted_tree <- read.tree("data/04_Phylogenetic_Tree/ASVs_unrooted.tree")
unrooted_tree
str(unrooted_tree)
```
#Merge Phyloseq Object
```{r merge-physeq}
# Intuition check 
stopifnot(ntaxa(raw_preprocessed_physeq) == ntaxa(unrooted_tree))

# Merge the tree with the phyloseq object 
unrooted_physeq <- 
  merge_phyloseq(raw_preprocessed_physeq, unrooted_tree)
unrooted_physeq
```

#Plot tree with 'ggtree'
```{r plot-tree-unrooted}
# Make a basic tree
kingdom_tree <- 
  ggtree(unrooted_physeq) + 
  # color tips by kingdom 
  geom_tippoint(mapping = aes(color = Kingdom)) + 
#  scale_color_manual(values = c("goldenrod1", "cornflowerblue", "grey")) +
  # Add title 
  labs(title = "Unrooted Tree") + 
  #move the legend to the bottom 
  theme(legend.position = "bottom");

kingdom_tree

kingdom_node_tree <- 
  kingdom_tree + 
  # Add the node label 
  geom_text(aes(label=node), hjust= -0.5, vjust = -0.3, size = 2)

kingdom_node_tree
```

Branch 1289 and branch 582 look suspiciously lengthy, but branch 1289 has many close ancestors while 582 is not as long so I would say that apart from that is relatively clean, the only other long clade is the archeal.

#Evaluating long branches
```{r evaluate-long-branch-582}
# Zoom in on node that looks potentially suspicious
viewClade(kingdom_node_tree + 
          labs(title = "Tree: Node 582"), 
          node = 582)
# Lets see its ancestors
ancestor(unrooted_tree, 582)
# Or if there are two nodes that we can see we can find its MRCA
## create tibble of tree to identify offspring
tree_tib <- as_tibble(unrooted_physeq@phy_tree)

# lets look at the branches
MRCA_2573 <- MRCA(unrooted_tree, .node1 = 582, .node2 = 583) #node = 2573

viewClade(kingdom_node_tree, 2573) 

## create dataframe to view ASVs and taxa info and to pull the ASVs
node2573 <- offspring(tree_tib, 2573, tiponly = TRUE) %>% 
  as.data.frame() %>% pull(label)

# Now, let's pull out the taxonomy information, including the ASV sequence!
node_2573_df <- 
  unrooted_physeq %>%
  tax_table() %>%
  as.data.frame() %>% 
  filter(ASV %in% node2573) 

# Take a look: 
glimpse(node_2573_df)

# look at counts of the ASVs
unrooted_physeq %>%
  subset_taxa(., ASV %in% node_2573_df$ASV) %>%
  otu_table() %>%
  data.frame() %>%
  colSums()
```

```{r evaluate-long-branch-2922}
#for node 2922
# Zoom in on node that looks potentially suspicious
viewClade(kingdom_node_tree+ 
          labs(title = "Tree: 2922"), 
          node = 2922)

# Lets see its ancestors
ancestor(unrooted_tree, 2922)
# Or if there are two nodes that we can see we can find its MRCA


# Let's evaluate 2921!!
viewClade(kingdom_node_tree, 2921) # even more zoomed out

# lets look at the long branch
mrca_node_2922 <- tidytree::MRCA(unrooted_tree, .node1 = 933, .node2 = 932) #2922

# Inspect
mrca_node_2922 #2922
## create dataframe to view ASVs and taxa info and to pull the ASVs
node_2922 <- 
  offspring(tree_tib, mrca_node_2922, tiponly = TRUE) %>% 
  as.data.frame() %>% 
  pull(label)

# Inspect
node_2922
# Now, let's pull out the taxonomy information, including the ASV sequence!
node_2922_df <- 
  unrooted_physeq %>%
  tax_table() %>%
  as.data.frame() %>% 
  filter(ASV %in% node_2922) 

# Take a look: 
glimpse(node_2922_df)

node_2922_df$ASVseqs
# look at counts of the ASVs
unrooted_physeq %>%
  subset_taxa(., ASV %in% node_2922_df$ASV) %>%
  otu_table() %>%
  data.frame() %>%
  colSums()
```

There were 2 relatively long branches in my tree but I did not remove neither, node 2573 only appeared twice in 1 sample but when doing fast blasting, the sequence was recognized at the genus level so I think it has biological significance.

#Root tree from archea
```{r root-archea}
# but first find the ASVs that are Archaea
archaea_node <- tax_table(raw_preprocessed_physeq) %>%
  as.data.frame() %>%
  filter(Kingdom == "Archaea") %>%
  pull(ASV)

# find the node that encompasses all Archaea
archaea_node <- findMRCA(unrooted_tree, tips = archaea_node, type = "node") #node = 2922

# View clade to check it out
viewClade(kingdom_node_tree, 2922)

# Root the tree using node 2922
archaeal_rooted_tree <- ape::root(unrooted_tree, node = 2922, resolve.root = TRUE)

# Check if tree is rooted
is.rooted(archaeal_rooted_tree)# should be TRUE

# plot unrooted tree with node 2922 highlighted
archaea_tree <- 
  ggtree(unrooted_physeq) +
  geom_tippoint(mapping = aes(color=Kingdom),
                size = 1.5) +
  geom_text(aes(label = node), hjust = -0.7, vjust = -0.5, size = 2) +
  geom_highlight(node = 2922, fill = "dodgerblue")

archaea_tree
# merge with physeq
archaeal_rooted_physeq <- merge_phyloseq(raw_preprocessed_physeq, archaeal_rooted_tree)

# Plot rooted simple tree with no text
archaea_rooted_simple <- ggtree(archaeal_rooted_physeq) +
  geom_tippoint(aes(color = Kingdom))  

archaea_rooted_simple

# Plot rooted tree with text
archaeal_rooted_text <- ggtree(archaeal_rooted_physeq) +
  geom_tippoint(aes(color = Kingdom)) + 
  geom_text(aes(label = node), hjust = -0.5, vjust = -0.3, size = 1.5)

archaeal_rooted_text
```
The tree looks finalized, the long branches have no reason to be pruned and overall the rest of the ASVs have close relationships to each other so there are no aditional points of interest.
#Save phyloseq object
```{r save-phyloseq}
# now lets save our preprocessed phyloseq object
save(archaeal_rooted_physeq, 
     file = "data/04_Phylogenetic_Tree/archaeal_rooted_physeq.RData")
```

#Session information
```{r session-info}
# Ensure reproducibility 
devtools::session_info()
```

