---
title: "Beta diversity of microbes during inulin treatment"
author: "Roger Calvas"
date: "`r Sys.Date()`"
output:
  html_document: 
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "../../figures/06_Ordination/")

```

# Goals 

1. Load in phyloseq data with rooted tree & colors and functions.  
2. Evaluate sequencing depth and remove samples, if needed.  
3. Normalize the read counts between samples.  
4. Calculate community **dis**similarities. Numbers between 0 and 1. If 0, completely similar versus if they are 1, then they're completely dissimilar.   
    a. **Sorensen**: Shared Species as a binary value: Abundance-unweighted 
    b. **Bray-Curtis**: Shared Abundant species: Abundance-weighted
    c. **(Abundance-)Weighted UNIFRAC**: Consider Abundant Species and where they fall on the tree  
5. Run statistics with functions from the vegan R package: 
    a. PERMANOVA with `adonis2()`.
    b. betadispR with `betadisper()` and `permutest()`. 
6. Visualize the community data with two unconstrained Ordinations:  
    a. **PCoA**: Linear Method. Eigenvalue = how much variation is explained by each axis. Choose to view axis 1, 2, 3, etc. and plot them together.  
    b. **NMDS**: Non-linear. Smush multiple Dimensions into 2 or 3 axes. Need to report Stress value (ideally <0.15).  

## Inputs 

1. We will need the `phytree_preprocessed_physeq.RData`

## Outputs 

*This is an exciting analysis!!! We get to answer another one of our scientific questions!*

1. Calculated beta-diversity dissimilarity measures (*e.g.* Sorensen, Bray-Curtis, abundance-unweighted and -weighted UniFrac) across every sample. 
2. Ordination figures (*i.e.,* PCoA/NMDS) to include in the scientific paper that visualize the data as it relates to the scientific question.
3. Statistical tests (*i.e.,* PERMANOVA & betadisper) conveying the measured and quantified changes and patterns in biodiversity.

# Scientific Question 

Does inulin treatment increase SCFA production (i.e., acetate, butyrate, propionate), and is this associated with a shift in fiber-fermenting bacteria?

- *Null Hypothesis:* Microbial diversity and scfa production does not vary during inulin treatment.
- *Alternative Hypothesis:* An inulin treatment will stimulate scfa production and shift microbial composition.


<span style="color: red;">INTERPRETATION #1:  What is your scientific question and hypotheses? Do they relate to abundance? Presence/absence? Phylogeny?</span>

Does inulin treatment increase SCFA production (i.e., acetate, butyrate, propionate), and is this associated with a shift in fiber-fermenting bacteria?

- *Null Hypothesis:* Microbial diversity and scfa production does not vary during inulin treatment.
- *Alternative Hypothesis:* An inulin treatment will stimulate scfa production and shift microbial composition.

This question relates to abundance by observing changes in the gut microbial composition during an inulin treatment, favoring the taxa that produce scfa from fermentable fiber.
# Set up 

## Timing of this script

```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
```

## Set the seed

```{r set-seed}
set.seed(39809473)
```

## Load Packages, colors & functions  
```{r load-packages}
pacman::p_load(tidyverse, devtools, phyloseq, patchwork, vegan, 
               install = FALSE)

# Load Colors 
source("code/colors.R")

# Load functions 
source("code/functions.R")
```

# 1. Load in the data 

```{r load-data}
# load phyloseq object
load("data/04_Phylogenetic_Tree/archaeal_rooted_physeq.RData")
archaeal_rooted_physeq

# Intuition check on seq depth
min(sample_sums(archaeal_rooted_physeq))
```

# Normalizing the Read Depth 

## 2. Explore the Raw Read Counts 

```{r explore-read-counts, fig.width=6, fig.height=3}
# calculate read counts per sample 
raw_TotalSeqs_df <- 
  archaeal_rooted_physeq %>%
  # Calculate the total number of sequences/reads
  sample_sums() %>%
  data.frame()

# Take a look 
head(raw_TotalSeqs_df)

# Rename the column 
colnames(raw_TotalSeqs_df)[1] <- "TotalSeqs"

# add a new column of num_ASVs (RAW, non-noramlized # ASVs)
raw_TotalSeqsASVs_df <- 
  raw_TotalSeqs_df %>%
  mutate(num_ASVs = colSums(otu_table(archaeal_rooted_physeq) > 1))

#View(raw_TotalSeqsASVs_df)

# Plot histogram of seq depth 
rawSeq_histogram <- 
  raw_TotalSeqsASVs_df %>%
  ggplot(aes(x = TotalSeqs)) + 
  geom_histogram(bins = 50) + 
  scale_x_continuous(limits = c(0, 35000)) + 
  labs(title = "Raw Seq Depth Histogram") + 
  theme_bw()

# Plot Seq Depth versus num_ASVs
rawSeq_vs_numASV_plot <- 
  raw_TotalSeqsASVs_df %>%
  ggplot(aes(x = num_ASVs, y = TotalSeqs)) + 
  geom_point() + 
  scale_y_continuous(limits = c(0, 35000)) +
  geom_smooth(method = "lm") + 
  labs(title = "Seq Depth vs # ASVs") + 
  theme_bw()

# is it significant? 
summary(lm(TotalSeqs ~ num_ASVs, data = raw_TotalSeqsASVs_df))

# Put the two plots together.
rawSeq_histogram + rawSeq_vs_numASV_plot +
  plot_annotation(tag_levels = "A")
```

<span style="color: red;">INTERPRETATION #2:  Evaluating the above plots, what did you scale the read counts to in your dataset? Why did you choose that number? Do you think it is sufficient to perform a beta-diversity analysis and accurately compare your data to one another?</span>

I chose the minimun sequencing depth in the dataset at 7k~ because even thought plot B shows a positive correlation between sequencing depth and number of ASVs but it is not too steep, also based on the spread of the scatterplot and the fact that the minimum is not an outlier it seems like sufficient lenght.
There is an outlier with unusually high count of ASVs so I will keep an eye on it.

# 3. Scale Read Counts 

```{r scale-reads, fig.width=6, fig.height=3}
min(sample_sums(archaeal_rooted_physeq))

# Scale the reads
scaled_physeq <- 
  archaeal_rooted_physeq %>%
  scale_reads(round = "matround")  

# Look at it 
scaled_physeq

# Save it
save(scaled_physeq, file ="data/06_Ordination/scaled_physeq.RData")

#Prune NA values for beta diversity analysis
scaled_physeq <-
  scaled_physeq %>%
  subset_samples(!is.na(total_scfa))

# Look at it more!
#View(data.frame(otu_table(archaeal_rooted_physeq)))
#View(data.frame(otu_table(scaled_physeq)))

# Confirm seq read depth of scaled_physeq 
scaled_TotalSeqs_df <- 
  scaled_physeq %>%
  sample_sums() %>%
  data.frame()

colnames(scaled_TotalSeqs_df)[1] <- "TotalSeqs"

# add a new column of num_ASVs (RAW, non-noramlized # ASVs)
scaled_TotalSeqsASVs_df <- 
  scaled_TotalSeqs_df %>%
  mutate(num_ASVs = colSums(otu_table(scaled_physeq) > 1))

# Plot it! 
scaledSeq_histogram <- 
  scaled_TotalSeqs_df %>%
  ggplot(aes(x = TotalSeqs)) + 
  geom_histogram(bins = 50) + 
  labs(title = "Scaled Seq Depth Histogram") + 
  scale_x_continuous(limits = c(0, 10000)) + 
  theme_bw()

# Scaling always gives us a range! 
# MInimum sequences
min(scaled_TotalSeqs_df$TotalSeqs) 
# Total Range 
range(scaled_TotalSeqs_df)
# How many is that range? 
range_seqs <- (max(scaled_TotalSeqs_df$TotalSeqs) - min(scaled_TotalSeqs_df$TotalSeqs))
range_seqs
# And in terms of percent? 
range_seqs/max(scaled_TotalSeqs_df$TotalSeqs)
#View(scaled_TotalSeqs_df)

# Set the axis min and max for the next plot 
max_y <- max(scaled_TotalSeqs_df$TotalSeqs) + 10
min_y <- min(scaled_TotalSeqs_df$TotalSeqs) - 10

# Now, draw the plot 
scaledSeq_vs_numASV_plot <- 
  scaled_TotalSeqsASVs_df %>%
  ggplot(aes(x = num_ASVs, y = TotalSeqs)) + 
  geom_point() + 
  scale_y_continuous(limits = c(min_y, max_y)) +
  #geom_smooth(method = "lm") + 
  theme_bw() + 
  labs(title = "Scaled: Seq Depth vs # ASVs")

# Put the two plots together.
scaledSeq_histogram + scaledSeq_vs_numASV_plot +
  plot_annotation(tag_levels = "A")


```

## Create metadata after pruning
```{r metadata}
# Create Metadata_df after the prune
metadata_df <- 
  scaled_physeq %>%
  sample_data() %>%
  data.frame()
```
<span style="color: red;">INTERPRETATION #3:  Evaluate the scaled read count plots above. Are the total number of sequences across all of your samples in the same general range? What is the minimum and maximum (i.e., the range) of the sequencing depth? How was the relationship with the number of ASVs modified after scaling the read counts?</span>

They are all in the same range, the minimum is 7191 and the maximum is 7229 ( range of 38), the range is 0.5~% which is really good. The relationship stayed quite similar after scaling, with the scatter plot distribution skewing a bit to the left including the outlier sample with the most ASVs.

##Calculate dissimilarity

```{r Calc-dissimilarity}
#Jaccard Dissimilarity
scaled_jaccard_dist <- phyloseq::distance(scaled_physeq, method = "jaccard", binary = TRUE)

# What does it look like? 
class(scaled_jaccard_dist)
str(scaled_jaccard_dist)
#head(as.matrix(scaled_jaccard_dist))

scaled_sorensen_dist <- phyloseq::distance(scaled_physeq, method = "bray", binary = TRUE)

# Bray-Curtis Dissimiliarty
scaled_Bray_dist <- phyloseq::distance(scaled_physeq, method = "bray", binary = FALSE)

# Abundance-Unweighted UniFrac
scaled_uUnifrac_dist <- phyloseq::distance(scaled_physeq, method = "unifrac")

# Abundance-Weighted UniFrac
scaled_wUnifrac_dist <- phyloseq::distance(scaled_physeq, method = "wunifrac")
```

<span style="color: red;">INTERPRETATION #4:  Which dissimilarity metric is most related to your scientific question? Why or why not?</span>

I think bray curtis would be the most related to my scientific question, since I am looking for shifts in the dinamics of the community I want to look closely into relative abundance changes during the treatment, also find specific taxa that increased in response to inulin and that are related to scfa production from fermentable fiber, unifrac is a close second because I want to see the shifts in the structure of the community.

### Permanova continous variable with total_scfa
```{r PERMANOVA-categorical-bray}
# bray
## 1. Run with by = terms for R² values, sensitive to order of variables! 
## ALWAYS check and confirm the order of your terms and how they interact with each other.  
bray_status_adonis_terms1 <- adonis2(scaled_Bray_dist ~ status + total_scfa, data = metadata_df, by = "terms")
bray_status_adonis_terms1

# Check the order of the terms
bray_status_adonis_terms2 <- adonis2(scaled_Bray_dist ~ total_scfa + status, data = metadata_df, by = "terms")
bray_status_adonis_terms2

## 2. Run with by = "margin" for marginal p-values, which we can compare to the residuals from the first one. 
bray_status_adonis_margin <- adonis2(scaled_Bray_dist ~ total_scfa * status, data = metadata_df, by = "margin")
bray_status_adonis_margin
```
### Permanova continous variable with total_scfa
```{r PERMANOVA-categorical-sorensen}
# sorensen
## 1. Run with by = terms for R² values, sensitive to order of variables! 
## ALWAYS check and confirm the order of your terms and how they interact with each other.  
sorensen_status_adonis_terms1 <- adonis2(scaled_sorensen_dist ~ status + total_scfa, data = metadata_df, by = "terms")
sorensen_status_adonis_terms1

# Check the order of the terms
sorensen_status_adonis_terms2 <- adonis2(scaled_sorensen_dist ~ total_scfa + status, data = metadata_df, by = "terms")
sorensen_status_adonis_terms2

## 2. Run with by = "margin" for marginal p-values, which we can compare to the residuals from the first one. 
sorensen_status_adonis_margin <- adonis2(scaled_sorensen_dist ~ total_scfa * status, data = metadata_df, by = "margin")
sorensen_status_adonis_margin
```
### Permanova continous for each scfa with bray
```{r different-scfa-permanova}
bray_status_adonis_propionate1 <- adonis2(scaled_Bray_dist ~ status + propionate, data = metadata_df, by = "terms")
bray_status_adonis_propionate1

# Check the order of the terms
bray_status_adonis_propionate2 <- adonis2(scaled_Bray_dist ~ propionate + status, data = metadata_df, by = "terms")
bray_status_adonis_propionate2

bray_status_adonis_Butyrate1 <- adonis2(scaled_Bray_dist ~ status + Butyrate, data = metadata_df, by = "terms")
bray_status_adonis_Butyrate1

# Check the order of the terms
bray_status_adonis_Butyrate2 <- adonis2(scaled_Bray_dist ~ Butyrate + status, data = metadata_df, by = "terms")
bray_status_adonis_Butyrate2

bray_status_adonis_acetate1 <- adonis2(scaled_Bray_dist ~ status + acetate, data = metadata_df, by = "terms")
bray_status_adonis_acetate1

# Check the order of the terms
bray_status_adonis_acetate2 <- adonis2(scaled_Bray_dist ~ acetate + status, data = metadata_df, by = "terms")
bray_status_adonis_acetate2
```

<span style="color: red;">INTERPRETATION #5: Using the above example to guide you, run the appropriate PERMANOVA for your scientific question and interpret the results. What biological conclusion can you draw from your PERMANOVA results?</span>

Across all models status explained 0.8% of the variation, so the inulin treatment may not be statistically significant but it may change the microbial composition in specific reproducible ways, aditionally all results had a high R^2 values meaning that the results may be explained by specific abundance changes in certain taxa. 
While status itself only explains a small portion of the variation scfa_production seems to be linked to the community shifts so it may be related to functional potential.

# 5b. Betadisper: Testing Variances/Dispersions
## Sorensen

```{r betadisper-sorensen}
# Homogeneity of Disperson test with beta dispr
# Sorensen Betadisper - status 
dispr_sorensen_status <- betadisper(scaled_sorensen_dist, metadata_df$status)
# permutest() performs a non-parametric permutation test, which is robust and valid for the kind of data used in beta diversity analysis (e.g., dissimilarity matrices).
permutest(dispr_sorensen_status)

# Sorensen Betadisper - propionate
dispr_sorensen_propionate<- betadisper(scaled_sorensen_dist, metadata_df$propionate)
permutest(dispr_sorensen_propionate)

# Sorensen Betadisper - Butyrate 
dispr_sorensen_Butyrate<- betadisper(scaled_sorensen_dist, metadata_df$Butyrate)
permutest(dispr_sorensen_Butyrate)
```

### Now lets test with bray curtis
```{r betadisper-bray}
# Bray-Curtis Betadisper - Status
dispr_bray_status<- betadisper(scaled_Bray_dist, metadata_df$status)
permutest(dispr_bray_status)

# Bray-Curtis Betadisper - propionate 
dispr_bray_propionate <- betadisper(scaled_Bray_dist, metadata_df$propionate)
permutest(dispr_bray_propionate)

# Bray-Curtis Betadisper - Butyrate  
dispr_bray_Butyrate <- betadisper(scaled_Bray_dist, metadata_df$Butyrate)
permutest(dispr_bray_Butyrate)
```


<span style="color: red;">INTERPRETATION #6: Using the above example to guide you, run the appropriate betadisper() and permutest() for your scientific question and interpret the results. What biological conclusion can you draw from your dispersion analysis?</span>

Based on the dispersion analysis I can conclude that only propionate has significant dispersion making propionate analysis less rigid, while all the other variables had not significant dispersion.


<span style="color: red;">INTERPRETATION #7: Taking into account both the betadisper()/permutest() and the PERMANOVA (adonis2()) test that you ran, what biological conclusion can you draw regarding your data?</span>

Given all the test results I can conclude that inulin has a small not-statistical significant effect on scfa concentrations, aditioanlly it has a relevant effect on microbial composition that can further explored.even if the permanova was not significant the difference is not explained by dispersion and it could account for true biological difference.

# PCoA: Principal Coordinate analysis
```{r pcoa-plots, fig.height=3.5, fig.width=7}
### SORENSEN 
# First, calculate PCoA with Soresen
scaled_soren_pcoa <- 
  ordinate(physeq = scaled_physeq,
         method = "PCoA",
         distance = "bray", binary = TRUE)

# Take a quick look
str(scaled_soren_pcoa)

#prework for triplot ( I am going to go back and re do a couple of steps considering that I am pruning the 51 NA samples instead of imputing, alpha diversity as well, for now I will finish the hw :())
###  Subset to samples with non-missing total_scfa
# Filter the scaled object to remove samples with NA total_scfa
physeq_no_na <- subset_samples(scaled_physeq, !is.na(total_scfa))
## subsetted distance object
subset_soren_dist <- distance(physeq_no_na, method = "bray", binary = TRUE)
## PCoA with vegan object
vegan_pcoa_soren <- capscale(subset_soren_dist ~ 1)
## Fit vector 
scfa_vector_fit_soren <- envfit(vegan_pcoa_soren, sample_data(physeq_no_na)$total_scfa, permutations = 999)
scfa_arrow_soren <- as.data.frame(scores(scfa_vector_fit_soren, display = "vectors"))
scfa_arrow_soren$variable <- "total_scfa"
# Rename to match the axis names in ordination_df_soren
colnames(scfa_arrow_soren)[1:2] <- c("CAP1", "CAP2")

## Prepare for plotting the vector 
ordination_df_soren <- as.data.frame(scores(vegan_pcoa_soren, display = "sites"))
colnames(ordination_df_soren)[1:2] <- c("CAP1", "CAP2")
ordination_df_soren$total_scfa <- sample_data(physeq_no_na)$total_scfa
ordination_df_soren$status <- sample_data(physeq_no_na)$status
ordination_df_soren$host_subject_id <- sample_data(physeq_no_na)$host_subject_id

# Variation explained per axis
## Extract eigenvalues from capscale ordination
eig_vals_soren <- vegan_pcoa_soren$CA$eig

## Calculate % variance explained by CAP1 and CAP2
var_exp_soren <- round(100 * eig_vals_soren / sum(eig_vals_soren), 1)

## Create axis labels
xlab_text_soren <- paste0("Axis 1 (", var_exp_soren[1], "%)")
ylab_text_soren <- paste0("Axis 2 (", var_exp_soren[2], "%)")

# Plot it: Sorensen PCoA  
sorensen_pcoa_plot <- 
  ggplot(ordination_df_soren, aes(x = CAP1, y = CAP2)) +
  geom_line(aes(group = host_subject_id), color = "gray80", linewidth = 0.5, alpha = 0.5) +
  geom_point(aes(color = total_scfa, shape = status), size = 3.5, alpha = 0.85, stroke = 0.2) +
  scale_color_gradientn(colors = c("grey80", "darkgreen", "green")) +
  scale_shape_manual(values = c(16,17))+
  geom_segment(data = scfa_arrow_soren, aes(x = 0, y = 0, xend = CAP1, yend = CAP2), arrow = arrow(length = unit(0.3, "cm"), type = "closed"), color = "lightgreen", linewidth = 1.5, alpha = 0.5) +
  geom_text(data = scfa_arrow_soren,
            aes(x = CAP1 -0.02, y = CAP2 -0.1, label = variable), color = "lightgreen", fontface = "bold", size = 3.5, alpha = 0.5) +
  labs(title = "Triplot: soren", x = xlab_text_soren, y = ylab_text_soren, color = "Total SCFA (mmol/kg)", shape = "status") +
  annotate("text",
           x = Inf, y = -Inf,
           label = "SCFA fit: R² = 0.017, p = 0.381",
           hjust = 1.1, vjust = -1.2,
           size = 3.5, fontface = "italic", color = "darkgreen") +
  theme_bw()

### Bray-Curtis 
# Second, calculate PCoA with Bray-Curtis
scaled_bray_pcoa <- 
  ordinate(physeq = scaled_physeq,
         method = "PCoA",
         distance = "bray", binary = FALSE)


## subsetted distance object
subset_bray_dist <- distance(physeq_no_na, method = "bray", binary = FALSE)
## PCoA with vegan object
vegan_pcoa_bray <- capscale(subset_bray_dist ~ 1)
## Fit vector 
scfa_vector_fit_bray <- envfit(vegan_pcoa_bray, sample_data(physeq_no_na)$total_scfa, permutations = 999)
scfa_arrow_bray <- as.data.frame(scores(scfa_vector_fit_bray, display = "vectors"))
scfa_arrow_bray$variable <- "total_scfa"
# Rename to match the axis names in ordination_df_bray
colnames(scfa_arrow_bray)[1:2] <- c("CAP1", "CAP2")

## Prepare for plotting the vector 
ordination_df_bray <- as.data.frame(scores(vegan_pcoa_bray, display = "sites"))
colnames(ordination_df_bray)[1:2] <- c("CAP1", "CAP2")
ordination_df_bray$total_scfa <- sample_data(physeq_no_na)$total_scfa
ordination_df_bray$status <- sample_data(physeq_no_na)$status
ordination_df_bray$host_subject_id <- sample_data(physeq_no_na)$host_subject_id

# Variation explained per axis
## Extract eigenvalues from capscale ordination
eig_vals_bray <- vegan_pcoa_bray$CA$eig

## Calculate % variance explained by CAP1 and CAP2
var_exp_bray <- round(100 * eig_vals_bray / sum(eig_vals_bray), 1)

## Create axis labels
xlab_text_bray <- paste0("Axis 1 (", var_exp_bray[1], "%)")
ylab_text_bray <- paste0("Axis 2 (", var_exp_bray[2], "%)")

# Plot it: braysen PCoA  
bray_pcoa_plot <- 
  ggplot(ordination_df_bray, aes(x = CAP1, y = CAP2)) +
  geom_line(aes(group = host_subject_id), color = "gray80", linewidth = 0.5, alpha = 0.5) +
  geom_point(aes(color = total_scfa, shape = status), size = 3.5, alpha = 0.85, stroke = 0.2) +
  scale_color_gradientn(colors = c("grey80", "darkgreen", "green")) +
  scale_shape_manual(values = c(16,17))+
  geom_segment(data = scfa_arrow_bray, aes(x = 0, y = 0, xend = CAP1, yend = CAP2), arrow = arrow(length = unit(0.3, "cm"), type = "closed"), color = "lightgreen", linewidth = 1.5, alpha = 0.5) +
  geom_text(data = scfa_arrow_bray,
            aes(x = CAP1 -0.02, y = CAP2 -0.1, label = variable), color = "lightgreen", fontface = "bold", size = 3.5, alpha = 0.5) +
  labs(title = "Triplot: bray", x = xlab_text_bray, y = ylab_text_bray, color = "Total SCFA (mmol/kg)", shape = "status") +
  theme_bw() +
    annotate("text",
           x = Inf, y = -Inf,
           label = "SCFA fit: R² = 0.028, p = 0.210",
           hjust = 1.1, vjust = -1.2,
           size = 3.5, fontface = "italic", color = "darkgreen")


print(scfa_vector_fit_soren)
print(scfa_vector_fit_bray)


# Show the plots 
 sorensen_pcoa_plot + bray_pcoa_plot +
   plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A")
```

<span style="color: red;">INTERPRETATION #8: Plot your data using a PCoA. Interpret what you can conclude from the PCoA.</span>
There is no strong clustering of samples by treatment with bray curtis clustering apparently because of scfa concentrations, and there is minimal correlation between community shifts and total_scfa levels, bray-curtis has a higher percentage explained in each axis and a stronger scfa vector, but still not statistically significant.
Neither reveal strong correlations between total SCFA production and community composition. This suggests that the small but consistent changes in each scfa are probably caused within individuals or by changes in functional potential.

## 6b. NMDS: Non-Metric Multidimensional Scaling 

```{r soren-nmds}
## SORENSEN 
scaled_soren_nmds <- 
  ordinate(physeq = scaled_physeq,
         method = "NMDS",
         distance = "bray", binary = TRUE)

# Plot it! 
sorensen_nmds_plot <- 
  plot_ordination(physeq = scaled_physeq,
                ordination = scaled_soren_nmds,
                color = "status",
                title = "Sorensen NMDS") + 
  scale_color_manual(values = status_colors) + 
  scale_shape_manual(values = c(15, 16, 17)) + 
  geom_point(size = 5, alpha = 0.5, aes(color = status)) + 
  theme_bw() + 
  theme(legend.position = "right")

### Bray-Curtis 
# Second, calculate NMDS with Bray-Curtis
scaled_bray_nmds <- 
  ordinate(physeq = scaled_physeq,
         method = "NMDS",
         distance = "bray", binary = FALSE)

# Plot it: Bray-Curtis NMDS 
bray_nmds_plot <- 
  plot_ordination(physeq = scaled_physeq,
                ordination = scaled_bray_nmds,
                color = "status",
                title = "Bray-Curtis NMDS") + 
  scale_color_manual(values = status_colors) + 
  scale_shape_manual(values = c(15, 16, 17)) + 
  geom_point(size = 5, alpha = 0.5, aes(color = status)) + 
  theme_bw() + 
  theme(legend.position = "right")

# Show the plots 
sorensen_nmds_plot + bray_nmds_plot + 
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A")
```


<span style="color: red;">INTERPRETATION #9: Plot your data using NMDS. Interpret what you can conclude from the NMDS.</span>

The stress is acceptable, just below the cutoff. But the graph itself shows no strong clustering effect by treatment status with a lot of overlapping.

## All ordinations together!

```{r ordinations, fig.width=7, fig.height=6}
sorensen_pcoa_plot + bray_pcoa_plot + 
sorensen_nmds_plot + bray_nmds_plot + 
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A")
```

<span style="color: red;">INTERPRETATION #10: Plot both the PCoA and NMDS together. Which of the two visualizations work best for your data? Explain your reasoning.</span>

Neither plot shows a strong correlation to the treatment groups, but PCoA has better potential at explaining the differences in the data with some trends and directionality observed. Bray curtis PCoA showed clustering around the top right with a noticeable gradient possibly related to SCFA levels, NMDS showd clustering but most samples were overlapping with no clear trends.

# Final info for Reproducibility 

## Check Render Time
```{r stop-time}
# Take the time now that we are at the end of the script
end_time <- Sys.time()
end_time 

# Echo the elapsed time
elapsed_time <- round((end_time - start_time), 3)
elapsed_time
```

## Session Information

```{r session-info}
# Ensure reproducibility with package version information
devtools::session_info()
```