---
title: "Assigning ASVs with DADA2"
author: "Roger Calvas"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center", 
                      fig.path = "../figures/02_AssignASVs/")
```
# Goals 

1. Infer an error model for in the filtered sequences, separately on forward and reverse reads.
2. Assign ASVs on both forward and reverse reads separately by applying the error model.
3. Merge forward and reverse ASVs into "contiguous ASVs".
4. Generate the first draft of ASV count table.
5. Quality Trimming of ASV lengths.
6. Remove chimeras. 
7. Assign Taxonomy with Silva Database. 
8. Write out relevant files: `asv_table`, `asvs_fasta`, `tax_table`, and `sample_data`.

## Input 

1. Filtered fastq files generated from `01_QualityTrimming.Rmd`.
2. Sample Name vector.

## Output 

1. ASV Count Table: `asv_table` (with and without sequence names)
2. ASV fasta file: `asvs_fasta` for building a phylogenetic tree at a later step.
3. Taxonomy Table  `tax_table`
4. Sample Information: `sample_data`  track the reads lots throughout DADA2 workflow. 

# Set up the Environment 

```{r set-seed}
# Set the seed for reproducibility
set.seed(39809473)

# Let's make a parameter to set the number of threads 
n_threads = 20
```

#Timing of the script
```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
```

## Load Packages 
```{r load-packages}
# Efficient package loading with pacman 
pacman::p_load(tidyverse, devtools, dada2, 
               patchwork, DT, install = FALSE)
```

## Load Filtered Fastq Files 

```{r load-filtered-fastqs}
# Place filtered seq files into a variable 
filtered_fastqs_path <- "data/01_DADA2/02_filtered_fastqs"

# Intuition check:
filtered_fastqs_path

# Create Forward vector 
filtered_forward_reads <- 
  list.files(filtered_fastqs_path, pattern = "R1_filtered.fastq.gz",
             full.names = TRUE)
# Check 
filtered_forward_reads[1:5]

# Reverse vector 
filtered_reverse_reads <- 
    list.files(filtered_fastqs_path, pattern = "R2_filtered.fastq.gz",
             full.names = TRUE)

# Check 
filtered_reverse_reads[1:5]
```

## Sample Names 
```{r sample-names}
# Create vector of sample names from the filenames 
sample_names <- sapply(strsplit(basename(filtered_forward_reads), "_"), `[`, 1) 

# Intuition Check 
head(sample_names)
```
## Learn the Errors
```{r learn-errors-MiSeq, fig.width=12, fig.height=8}
# Forward Reads 
error_forward_reads <- 
  learnErrors(filtered_forward_reads, multithread = n_threads)

# Forward Error Plot 
forward_error_plot <- 
  plotErrors(error_forward_reads, nominalQ = TRUE) + 
  labs(title = "Forward Reads: Error Model")

# Reverse Reads 
error_reverse_reads <- 
  learnErrors(filtered_reverse_reads, multithread = n_threads)

# Reverse Error Plot 
reverse_error_plot <- 
  plotErrors(error_reverse_reads, nominalQ = TRUE) + 
  labs(title = "Reverse Reads: Error Model")

# Look at the plots together 
forward_error_plot + reverse_error_plot
```

#Infer ASVs
```{r Infer-ASVs}
# Infer ASVs on the forward sequences
dada_forward <- 
  dada(filtered_forward_reads, 
       err = error_forward_reads,
       multithread = n_threads) 
# Take a look at the data
# What type of data structure is it? 
typeof(dada_forward) # It's a list 
length(dada_forward) # How big is it? One per sample!

# What doees it look like for each sample?  
dada_forward$`SRR6445038_R1_filtered.fastq.gz`

# Reverse ASVs
dada_reverse <- 
  dada(filtered_reverse_reads,
       err = error_reverse_reads ,
       multithread = n_threads)

# Check data 
dada_reverse[30]
```
# Merge Forward and Reverse ASVs 
```{r merge-ASVs}
merged_ASVs <- 
  mergePairs(dada_forward, filtered_forward_reads,
             dada_reverse, filtered_reverse_reads,
             verbose = TRUE)

# Evaluate the data output 
typeof(merged_ASVs) # A list
length(merged_ASVs) # Length of the number of samples!
head(names(merged_ASVs)) # Here, we can access our current sample names

# Inspect further for each sample
#head(merged_ASVs, n = 2) # A dataframe for each sample
# We have a dataframe in each part of our list! What are in the columns? 
glimpse(merged_ASVs$`SRR6445038_R1_filtered.fastq.gz`)

summary(sapply(merged_ASVs, nrow))  # Number of merged reads per sample

```

# Create Raw ASV Count Table 

```{r raw-ASV-count-table}
# Raw ASV
raw_ASV_table <- makeSequenceTable(merged_ASVs)

# Intuition Check: Type and dimensions of the data
dim(raw_ASV_table)
typeof(raw_ASV_table)
class(raw_ASV_table)

# write out raw_asv_table 
write.table(raw_ASV_table, file = "data/01_DADA2/raw_ASV_counts.tsv",
            sep = "\t", quote = FALSE, col.names = NA)
```
1. What is the sequencing strategy:
Paired-end 250bp (2×250bp) Illumina MiSeq sequencing, targeting the V4 region with 515F-806R primers.
2. What’s the total length of our starting amplicons?
806 -515 = 291
total length with primers
3. What is the ASV length without primers?

The 515F (GTGYCAGCMGCCGCGGTAA) primer is 19 base pairs.
The 806R (GGACTACNVGGGTWTCTAAT) primer is 20 base pairs.
291 - 19 - 20 = 252

4.What is the length of the trimmed ASVs after filterAndTrim()?
In our 01_QualityTrimming.Rmd, we used trimLeft = 8 to remove the bases at the beginning of the forward read and truncLen = c(250, 248), which kept the full forward read length and removed 2 bases at the end of the reverse read. Therefore:
252 - 8 = 244
5.What is the overlap between our forward and reverse reads?
We performed 2x250 paired-end Illumina MiSeq Sequencing. And the multiqc reports a sequence length of 250.
Because of the unusable dips in the reverse reads after 200~ bases we had to use truncLen = c(240, 200), as to not lose too much sequence depth, because of this we are expected to get somewhere around 200 bp overlap.

### ASV Length Stats 

```{r assess-ASV-length}
# Calculate summary stats
# Longest ASV?
maxLength_ASV <- max(nchar(getSequences(raw_ASV_table)))

# Shortest ASV?
minLength_ASV <- min(nchar(getSequences(raw_ASV_table))) 

# Mean ASV length?
meanLength_ASV <- mean(nchar(getSequences(raw_ASV_table))) 

# Median ASV length?
medianLength_ASV <- median(nchar(getSequences(raw_ASV_table))) 

# Create a table to Inspect the distribution of sequence lengths of all ASVs in dataset 
table(nchar(getSequences(raw_ASV_table)))
```

### ASV Length Plot 
```{r ASV-lengths-plot}
# Inspect the distribution of sequence lengths of all ASVs in data set 
# AFTER TRIM
plot_ASVLength_raw <- 
  data.frame(Seq_Length = nchar(getSequences(raw_ASV_table))) %>%
  ggplot(aes(x = Seq_Length )) + 
  geom_histogram() + 
  # include the x-axis scales
  scale_x_continuous(limits = c(0, maxLength_ASV + 5)) + 
  labs(y = "Number of ASVs", x = "Raw ASV Length (bps)")

# Show the plot
plot_ASVLength_raw
```
I am surprised that the ASV lengths are at around 237 because I used trunclen (240, 200) and I tought that eliminating 50 bp on the reverse meant that I was gonna get lengths of 200 as well. I am pretty pleased with the sequence length given that I was more focused on mantaining sequence depth.
-----
My sequences are mainly on the 237 and 245 bp but since in the 16s v4 region we dont expect much variation of length then I am choosing the 237 bp which has more ASVs, since 237-245 would be too big of a range with 2 peaks on opposite sides.

## What are the current ASV lengths? 

```{r trim-ASVs}
# Subset only ASVs that are 245 bps long 
raw_ASV_table_trimmed <- 
  raw_ASV_table[, nchar(colnames(raw_ASV_table)) >= 244 & nchar(colnames(raw_ASV_table)) <= 245]

# Inspect the distribution of sequence lengths of all ASVs in dataset 
table(nchar(getSequences(raw_ASV_table_trimmed)))

# What proportion of total ASV sequences are left in the data? 
percRetained_Trimmed <- sum(raw_ASV_table_trimmed)/sum(raw_ASV_table)
percRetained_Trimmed # Show it 

# Inspect the distribution of sequence lengths of all ASVs in dataset 
# AFTER TRIM
plot_ASVLength_trimmed <- 
  data.frame(Seq_Length = nchar(getSequences(raw_ASV_table_trimmed))) %>%
  ggplot(aes(x = Seq_Length )) + 
  geom_histogram() + 
  # include the x-axis scales
  scale_x_continuous(limits = c(0, maxLength_ASV + 5)) + 
  labs(y = "Number of ASVs", x = "Trimmed ASV Length (bps)")

# Show the plot 
plot_ASVLength_trimmed
```

# Remove Chimeras
```{r rm-chimeras, fig.width=3.5, fig.height=3}
# Remove the chimeras in the raw ASV table
noChimeras_ASV_table <- 
  removeBimeraDenovo(raw_ASV_table_trimmed, 
                     method="consensus", 
                     multithread = n_threads, 
                     verbose=TRUE)

# Check the dimensions
dim(noChimeras_ASV_table)

# What proportion is left of the sequences? 
# Chimera removal compared to trimming 
percRetained_chimerasTrimmed <- sum(noChimeras_ASV_table)/sum(raw_ASV_table_trimmed)
# Chimera removal compared to raw  
percRetained_chimerasRaw <-sum(noChimeras_ASV_table)/sum(raw_ASV_table)

# Plot it 
plot_ASVLength_NoChimeras <- 
  data.frame(Seq_Length_NoChim = nchar(getSequences(noChimeras_ASV_table))) %>%
  ggplot(aes(x = Seq_Length_NoChim )) + 
  geom_histogram()+ 
  # include the x-axis scales
  scale_x_continuous(limits = c(0, maxLength_ASV + 5)) + 
  labs(y = "Number of ASVs \n (Post-Chimera Removal)", 
       x = "ASV Length (bps)")

# Show the plot
plot_ASVLength_NoChimeras 
```

1. `r round(percRetained_Trimmed * 100, digits = 2)`% of the raw ASVs were retained *after trimming*.
2. `r round(percRetained_chimerasTrimmed *100, digits = 2)`% of the trimmed ASV counts were retained *after chimera removal*.
3. This translates to `r round(percRetained_chimerasRaw *100, digits=2)`% retention of the original, raw merged ASV counts after *both trimming and chimera removal.*


### Plot ASV Lengths 

```{r plot-ASVLengths, fig.height=2.5, fig.width=7}
plot_ASVLength_raw + plot_ASVLength_trimmed + plot_ASVLength_NoChimeras + 
    plot_annotation(tag_levels = 'A')
```

# Track the read counts
```{r track-reads, fig.width=6, fig.height=4}
# A little function to identify number seqs 
getN <- function(x) sum(getUniques(x))

# Make the table to track the seqs 
track <- cbind(sapply(dada_forward, getN),
               sapply(dada_reverse, getN),
               sapply(merged_ASVs, getN),
               rowSums(noChimeras_ASV_table))

head(track)

# Update column names to be more informative (most are missing at the moment!)
colnames(track) <- c("denoisedF", "denoisedR", "merged", "nochim")
rownames(track) <- row.names(noChimeras_ASV_table)

# Generate a dataframe to track the reads through our DADA2 pipeline
track_counts_df <- 
  track %>%
  # make it a dataframe
  as.data.frame() %>%
  rownames_to_column(var = "sample_names")

# Now let's add a column for the number of ASVs
# First, intuition check that the samples match 
stopifnot(track_counts_df$sample_names == row.names(noChimeras_ASV_table))

# Now, let's add a new column with the number of ASVs
track_counts_df <- 
  track_counts_df %>%
  mutate(num_ASVs = rowSums(noChimeras_ASV_table > 1))

# Visualize it in table format 
DT::datatable(track_counts_df)

# Plot it!
track_counts_df %>%
  pivot_longer(denoisedF:nochim, names_to = "read_type", values_to = "num_reads") %>%
  mutate(read_type = fct_relevel(read_type, "denoisedF", "denoisedR", "merged", "nochim")) %>%
  ggplot(aes(x = read_type, y = num_reads, fill = read_type)) + 
  geom_line(aes(group = sample_names), color = "grey") + 
  geom_point(shape = 21, size = 3, alpha = 0.8) + 
  scale_fill_brewer(palette = "Spectral") + 
  labs(x = "Filtering Step", y = "Number of Sequences") + 
  theme_bw()
```

Now, let's plot the number of sequences that have been maintained in our samples using `geom_histogram()`. 
The proportion of raw ASVs retained is quite low, this may indicate effective trimming of chimeras. The final ASVs do infact cluster around the expected length of 245 so it was appropiate to choose that bp site.
```{r numSeqsASV-plot, fig.height=2.5, fig.width=7}
plot_ReadDepth <- 
  track_counts_df %>%
  ggplot(aes(x = nochim)) + 
  geom_histogram() + 
  labs(x = "Total # of Sequences", y = "# of Samples") + 
  theme_bw()
plot_ReadDepth
# What is the ASV richness per sample? 
plot_ASVRichness <- 
  track_counts_df %>%
  ggplot(aes(x = num_ASVs)) + 
  geom_histogram() + 
  labs(x = "Total # of ASVs", y = "# of Samples") + 
  theme_bw()
plot_ASVRichness
# Now, let's look at the relationship of ASVs and Sequencing depth 
plot_ReadDepth_ASVRichness <- 
  track_counts_df %>%
  ggplot(aes(x = nochim, y = num_ASVs)) + 
  geom_point() + 
  labs(x = "Total # of Sequences", y = "# of ASVs") + 
  theme_bw()
plot_ReadDepth_ASVRichness
# Show the plots together 
plot_ReadDepth + plot_ASVRichness + plot_ReadDepth_ASVRichness + 
    plot_annotation(tag_levels = 'A')
```
The first plot is a histogram that shows the distribution of sequences cross samples, most of the samples have very few sequences, a smaller subset has a lot of sequences, this could indicate low quality or over trimming.
the second histogram detects the ASVs per sample, there is a similar trend to plot A, with some aditional information about the depth across samples.
The third is a scatter plot that shows the relationship between sequence number and ASVs number per sample, there is a positive slope indicating that more sequences lead to more ASVs, the spread shows that different samples have different microbial diversity as expected.

## How does the taxonomy work? 

The `assignTaxonomy` function  implements the Ribosomal Database Project (RDP) Naive Bayesian Classifier algorithm described in Wang et al. (2007), published in Applied and Environmental Microbiology, with a kmer size of 8 and 100 bootstrap replicates. 

```{r assign-tax}
# Assign up to genus level 
taxa_train <- 
  assignTaxonomy(noChimeras_ASV_table, 
                 refFasta = "/workdir/in_class_data/taxonomy/silva_nr99_v138.2_toGenus_trainset.fa.gz", 
                 multithread = n_threads)

# Add the genus/species information 
taxa_addSpecies <- 
  addSpecies(taxa_train, 
              refFasta = "/workdir/in_class_data/taxonomy/silva_v138.2_assignSpecies.fa.gz")

# Inspect the taxonomy 
glimpse(taxa_addSpecies) # Note that the rownames are the ASV sequences!
# Let's removing the ASV sequence rownames for display only
taxa_print <- taxa_addSpecies 
rownames(taxa_print) <- NULL
head(taxa_print)
#View(taxa_print)
```


# Export the Data

Remember from the beginning of the file under "Goals", we wanted to export 4 different types of information: 

1. ASV Count Table: `asv_table` (with and without sequence names)
2. ASV fasta file: `asvs_fasta` for building a phylogenetic tree at a later step.
3. Taxonomy Table  `tax_table`
4. Sample Information: `sample_data`  track the reads lots throughout DADA2 workflow. 


## 1. ASV Tables

We will export the following two ASV count tables, which will be in "long" format where the **ASVs are in rows** and the **sample names are in the columns.**

1. **With ASV seqs:** ASV headers include the *entire* ASV sequence 245 bases.
2. **with ASV names:** This includes re-written and shortened headers like ASV_1, ASV_2, etc, which will match the names in our fasta file below.

### Structure of ASV tables

Let's check that the ASVs are in the rows and the sample names are in the columns. If not, we can use the transpose function `t()` in R. 

```{r structure-ASV-table}
# What's the current format of the ASV table?
head(rownames(noChimeras_ASV_table)) # Samples!
head(colnames(noChimeras_ASV_table)) # ASV Sequences

# Therefore, we need to transpose the matrix 
final_ASV_table_withSeqs <- t(noChimeras_ASV_table)

# Intuition check
head(rownames(final_ASV_table_withSeqs)) # ASV Sequences
head(colnames(final_ASV_table_withSeqs)) # Sample names
```

### Names in ASV tables

#### Fix Sample Names 

Notice above that the sample names are actually file names. Let's fix this! 

> WARNING! This is an easy place where a silent error can step in where we could misname our samples, leading to catastrophe down the road. So, we need to be very careful and complete our checks and balances so we ensure the samples are not mixed up in the renaming process. 

```{r SampleNames-ASV-table}
# Remember at the top of the file we created a vector of sample names 
head(sample_names)
# Let's check with the actual column names 
head(colnames(final_ASV_table_withSeqs)) # Sample names
# And then apply our sample name script to check, too
head(sapply(strsplit(colnames(final_ASV_table_withSeqs), "_"), `[`,1)) # Looks good! 

# Now, add a break in the script break if this isn't true! 
# Let's make sure the sample names match the file names in the matrix.
stopifnot(sapply(strsplit(colnames(final_ASV_table_withSeqs), "_"), `[`,1) == sample_names)

# Now, we've done some checks to prove to ourselves there will be no silent errors, 
# Let's rename! 
colnames(final_ASV_table_withSeqs) <- sample_names
head(colnames(final_ASV_table_withSeqs))
```

#### Rename ASVs

Then, we can also fix the names of our ASVs for our second ASV table where we will replace the ASV sequence names with ASV_1, ASV_2

```{r prepare-ASVcount-Seqtable}
# Give headers more manageable names
# First pull the ASV sequences from the rownames
ASV_seqs <- rownames(final_ASV_table_withSeqs)
ASV_seqs[1:5]

# How many ASVs? 
num_ASVs <- dim(final_ASV_table_withSeqs)[1] # select the number of rows
num_ASVs 

# Make an empty vector the length of the number of ASVs, 
# which is where we will place the new operational ASV names 
ASV_headers <- vector(num_ASVs, mode = "character")

# Let's mae sure we have an empty vector!
ASV_headers[1:5]
length(ASV_headers) # looks good! 

# Now, let's create a vector with ASV numbers
# loop through vector and fill it in with ASV names 
for (i in 1:num_ASVs) {
  # Add leading zero to ASV name so they print in correct order.
  ASV_number <- sprintf("%04d", i)
  # Now, rename each spot in the ASV header vector as we loop through the for loop
  ASV_headers[i] <- paste(">ASV", ASV_number, sep = "_")
}

# Intuition check
ASV_headers[1:5]

# Create a new ASV table, which will have the ASV numbers as names 
# View(noChimeras_ASV_table) # To view the table
final_ASV_table <- final_ASV_table_withSeqs
glimpse(final_ASV_table)

## Replace the ASV seqs with the ASV numbers 
row.names(final_ASV_table) <- sub(">", "", ASV_headers)
final_ASV_table[1:5, 1:5]
View(final_ASV_table) # To view the table
``` 

### Write the ASV Tables!

1. **With ASV seqs:** ASV headers include the *entire* ASV sequence 245 bases.
2. **with ASV names:** This includes re-written and shortened headers like ASV_1, ASV_2, etc, which will match the names in our fasta file below.

```{r write-asv-tables}
# 1. Write count table with ASV sequence names
write.table(final_ASV_table_withSeqs, 
            file = "data/01_DADA2/ASV_table_withSeqNames.tsv", 
            sep = "\t", quote = FALSE, col.names = NA)

# 2. Write count table with ASV numbered names (e.g. ASV_1, ASV_2, etc)
write.table(final_ASV_table, 
            file = "data/01_DADA2/ASV_table.tsv", 
            sep = "\t", quote = FALSE, col.names = NA)
```


## 2. ASV Fasta File 

Now, let's create a fasta file that has our ASV numbers with each of their corresponding ASV sequences. This file needs to be created because we will use it later to build a phylogenetic tree. 

**Why do we need a phylogenetic tree?** Many widely used microbial ecology metrics rely on phylogenetic relationships to quantify community similarity. For example, there are phylogenetic alpha-diversity metrics like the [phylogenetic Hill Numbers](https://www.annualreviews.org/content/journals/10.1146/annurev-ecolsys-120213-091540) and also beta-diversity measures that incorporate the branch lenghts of phylogenetic trees as it relates to community dissimimlarity, [like the abundance-unweighted and abundance-weighted UniFrac dissimilarity metrics](https://journals.asm.org/doi/full/10.1128/aem.71.12.8228-8235.2005). These metrics consider species presence and their evolutionary distances, conveying relationships of communities through an ecological *and* evolutionary framework. 

### Write the ASV Fasta File

```{r write-asv-fastas}
# Let's take our asv_headers
head(ASV_headers, n = 2)
# And combine it with their sequences
head(ASV_seqs, n = 2)

# Combine in a fasta format with the cbind() function
ASV_fasta <- c(rbind(ASV_headers, ASV_seqs))
head(ASV_fasta, n = 4)

# Then, let's write it to a fasta file!
# This will be our reference later on for which seq matches which ASV
write(ASV_fasta, "data/01_DADA2/ASVs.fasta")
```

## 3. Taxonomy Table 

Before we actually write the taxonomy table, let's put the ASV sequences as another column in our taxonomy table and then replace the rownames to be our numbered ASV names. 

### Reformat Taxonomy 

1. Move ASV sequences to column 
2. Replace ASV rownames with ASV numbered names

> WARNING! This is another easy place where a silent error can step in!! Here, we could accidentally misname our ASVs. So, we need to be very careful and perform intuition checks to ensure that the ASV names are not mixed up in the renaming process. 

```{r reformat-tax-table}
# Inspect the taxonomy table
dim(taxa_addSpecies) # ASVs are in rows and Kingdom, Phylum, etc in Columns 
colnames(taxa_addSpecies) # Column names are Linnean Taxonomy 
head(rownames(taxa_addSpecies), n = 2) # ASV names are rownames 
class(taxa_addSpecies) # Character matrix

##### Prepare tax table 
# 1. Add the ASV sequences from the rownames to a column 
new_tax_table <- 
  taxa_addSpecies%>%
  as.data.frame() %>%
  rownames_to_column(var = "ASVseqs") 

# Intuition check 
glimpse(new_tax_table)

# IMPORTANT! Let's do our intuition check 
# This is where we ensure we don't mix up the ASV names!
stopifnot(new_tax_table$ASVseqs == rownames(final_ASV_table_withSeqs))

# Now let's add the ASV names 
rownames(new_tax_table) <- rownames(final_ASV_table)
head(new_tax_table)

### Final prep of tax table. Add new column with ASV names 
ASV_tax_table <- 
  new_tax_table %>%
  # add rownames from count table for phyloseq handoff
  mutate(ASV = rownames(final_ASV_table)) %>%
  # Reorder the columns
  dplyr::select(Kingdom, Phylum, Class, Order, Family, Genus, Species, ASV, ASVseqs)

# Assign the rownames, which is required by phyloseq
rownames(ASV_tax_table) <- ASV_tax_table$ASV

# Take a quick look
glimpse(ASV_tax_table)

# Intution check
stopifnot(ASV_tax_table$ASV == rownames(ASV_tax_table), 
          rownames(ASV_tax_table) == rownames(ASV_tax_table))
```

### Write the Taxonomy Table 

Now, let's write out the taxonomy table 

```{r write-tax-table}
# Write the table 
write.table(ASV_tax_table, 
            file = "data/01_DADA2/ASV_taxonomy.tsv", 
            sep = "\t", quote = FALSE, col.names = NA)
```


## 4. Sample Data 

Let's save the `track_counts_df`, which can be useful for downstream analyses. 

```{r save-track reads}
# And save the track_counts_df a R object, which we will merge with metadata information in the next step of the analysis in analysis/02_Taxonomic_Assignment. 
save(track_counts_df, file = "data/01_DADA2/track_read_counts.RData")
```

# Final info for Reproducibility 

## Check Render Time
```{r stop-time}
# Take the time now that we are at the end of the script
end_time <- Sys.time()
end_time 

# Echo the elapsed time
elapsed_time <- round((end_time - start_time), 3)
elapsed_time
```

## Session Information

```{r session-info}
# Ensure reproducibility with package version information
devtools::session_info()
```