---
title: "Biodiversity accross a fiber treatment"
author: "Roger Calvas"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.path = "../../figures/05_Biodiversity/")
```

# Goals 

Run an analysis on the within-sample biodiversity of the samples in our project. 

1. Calculate the Hill Numbers of the samples using the iNEXT package. 
2. Plot and evaluate the interpolated and extrapolated rarefaction curves from iNEXT. 
3. Analyze the Hill numbers as it relates to stations and salinity (in PSU).
4. Test and then answer the scientific question at hand! (Remember, we're doing biology!)

## Inputs 

1. We will need the `phytree_preprocessed_physeq`, which includes a rooted tree (ideally within the archaea!) that we created in `analysis/04B_Phylogenetic_Tree_Inspection.Rmd`. 

## Outputs 

*This is an exciting analysis!!! We finally get to answer one of our scientific questions!*

1. Calculated biodiversity measures (Hill numbers) across every sample. 
2. Figures to include in our scientific paper that visualize the data as it relates to the scientific question.
3. Statistical tests conveying the measured and quantified changes and patterns in biodiversity.

# Scientific Question

## Question & Hypotheses

How is produciton of short chain fatty acid modified in the gut microbiome after a dietary intervention with inulin?

- *Null Hypothesis:* Concentrations of scf does not change after a treatment with inulin.
- *Alternative Hypothesis:* Concentrations of scf in the gut increases after a treatment with inulin.

# Set up the Environment 

## Timing of this script

```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
```

## Set the seed 
```{r set-seed}
set.seed(3980)
```

## Load Packages & Colors 

```{r load-packages-colors}
# install packages for stats
#install.packages("rstatix")
#install.packages("ggpubr")

pacman::p_load(tidyverse, devtools, patchwork, iNEXT, phyloseq,
               # packages for stats
               ggpubr, rstatix, install = FALSE)

# Load in colors for plotting 
source("code/colors.R")
```

## Load in Data 

```{r load-data}
load("data/04_Phylogenetic_Tree/archaeal_rooted_physeq.RData")

# take a look at it! 
archaeal_rooted_physeq

# Intuition check on seq depth
min(sample_sums(archaeal_rooted_physeq))

# Make a metadata dataframe 
metadata_df <- 
  archaeal_rooted_physeq %>%
  sample_data() %>%
  data.frame()

# view of it
glimpse(metadata_df)
```

## Run iNEXT

```{r iNEXT-biodiversity}
# prepare the input data 
# Species in rows and the samples in columns 
iNEXT_input_df <- 
  archaeal_rooted_physeq %>%
  otu_table() %>%
  data.frame()

# Inpect 
dim(iNEXT_input_df)
iNEXT_input_df[1:5, 1:5]

# Run iNEXT 
# Remember to set the seed!
# ASVs in ROWs, samples in COLUMNS 
#iNEXT_data <- iNEXT(iNEXT_input_df, 
 #                   q = c(0, 1, 2),
 #                   datatype = "abundance")
load("data/05_Biodiversity/iNEXT_data.RData")
# Inspect 
str(iNEXT_data)
typeof(iNEXT_data)

#save(iNEXT_data, file = "data/05_Biodiversity/iNEXT_data.RData")
```

# Rarefaction Curves 

## `ggiNEXT` Rarefaction Curves 

```{r iNEXT-colors}
# Prepare Colors 
color_df <- 
  iNEXT_input_df %>%
  colnames() %>%
  data.frame()

# Check
head(color_df)

# Rename the column 
colnames(color_df)[1] <- "names"
# Check
head(color_df)

# Rename the column 
colnames(color_df)[1] <- "names"
# Check
head(color_df)

# Make a helper dataframe for plotting with colors 
iNEXT_color_df <- 
  color_df %>%
  # Merge with metadata
  left_join(metadata_df, by = "names") %>%
  # Merge with colors for plotting with ggiNEXT
  left_join(data.frame(status_colors = status_colors,
            status = names(status_colors)),
            by = "status")

head(iNEXT_color_df)
```

```{r plot-inext-rarefaction}
ggiNEXT(iNEXT_data, type = 1, facet.var = "Order.q") + 
  scale_color_manual(values = iNEXT_color_df$status_colors) + 
  scale_fill_manual(values = iNEXT_color_df$status_colors) + 
  scale_shape_manual(values = base::rep(17, nsamples(archaeal_rooted_physeq))) + 
  theme_bw() + 
  theme(legend.position = "none") + 
  labs(y = "Effective Number of ASVs", x = "Number of Sequences") 
```
## Manual Rarefaction Curves 

```{r manual-rarefaction}
str(iNEXT_data)

iNEXT_manual_df <- 
  iNEXT_data$iNextEst$size_based %>%
  dplyr::rename(names = Assemblage) %>%
  # join with the metadata
  left_join(., metadata_df, by = "names") %>%
  mutate(status = factor(status, levels = c("before",
                                              "during")))

# Inspection
dim(iNEXT_manual_df)

# Manually plot rarefaction curves 
iNEXT_manual_df %>%
  # Filter out rows that are calculated with actual sequences (no extrapolated)
  dplyr::filter(Method == "Rarefaction") %>%
  # Now, let's draw the plot! 
  ggplot(aes(x = m, y = qD, color = status, group = names)) + 
  geom_line() + 
  # Facet by station 
  facet_grid(Order.q~status, scales = "free") + 
  scale_color_manual(values = status_colors) + 
  theme_bw() + 
  labs(y = "Effective Number of ASVs", x = "Number of Sequences") + 
  theme(legend.position = "none")
```
#Interpretation 1:
Sequencing depth is likely sufficient, all curves plateau even in richness. There is an outlier sample during the treatment that has an unusually high amount of ASVs, I will need to be in the lookout for that.

*Run Shapiro-Wilk for the scf*

```{r shapiro-wilk-scf}

# Test of the data is normal for the continuous value of butyrate
shapiro.test(metadata_df$Butyrate)
# Test of the data is normal for the continuous value of acetate
shapiro.test(metadata_df$acetate)
# Test of the data is normal for the continuous value of butyrate
shapiro.test(metadata_df$propionate)
# Test of the data is normal for the continuous value of butyrate
shapiro.test(metadata_df$total_scfa)
```
Now for the concentration of the 3 scf and total scfa, all of them had p-value<0.05 so none of them are normal, now we can look into richness
```{r shapiro-wilk-richness}
### Now what about for richness? 
obs_div_df <- 
  iNEXT_manual_df %>%
  dplyr::filter(Method == "Observed") 

# check it
glimpse(obs_div_df)

# Pull out unique data from the three fractions of samples 
obs_whole_rich_df <- 
  obs_div_df %>%
  dplyr::filter(Order.q == 0)

# Test of the data is normal for the continuous value of richness
shapiro.test(obs_whole_rich_df$qD)
```
Now richness is not normal, so we need to use non parametric tests.

Kruskal wallis for the different scf
```{r scf, fig.height=4, fig.width=4}
# Calculate the kruskall-wallis stats
kw_status_butyrate <- 
  obs_whole_rich_df %>%
  kruskal_test(Butyrate ~ status) 

# Look at it 
kw_status_butyrate

# Calculate the kruskall-wallis stats
kw_status_acetate <- 
  obs_whole_rich_df %>%
  kruskal_test(acetate ~ status) 

# Look at it 
kw_status_acetate

# Calculate the kruskall-wallis stats
kw_status_propionate <- 
  obs_whole_rich_df %>%
  kruskal_test(propionate ~ status) 

# Look at it 
kw_status_propionate

# Calculate the kruskall-wallis stats
kw_status_total_scfa <- 
  obs_whole_rich_df %>%
  kruskal_test(total_scfa ~ status) 

# Look at it 
kw_status_total_scfa

```