---
title: "Biodiversity accross a fiber treatment"
author: "Roger Calvas"
date: "`r Sys.Date()`"
output: 
  html_document: 
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.path = "../../figures/05_Biodiversity/")
```

# Goals 

Run an analysis on the within-sample biodiversity of the samples in our project. 

1. Calculate the Hill Numbers of the samples using the iNEXT package. 
2. Plot and evaluate the interpolated and extrapolated rarefaction curves from iNEXT. 
3. Analyze the Hill numbers as it relates to status and short chain fatty acids concenctrations in milimoles per kilogram.
4. Test and then answer the scientific question at hand! (Remember, we're doing biology!)

## Inputs 

1. We will need the `phytree_preprocessed_physeq`, which includes a rooted tree (ideally within the archaea!) that we created in `analysis/04B_Phylogenetic_Tree_Inspection.Rmd`. 

## Outputs 

*This is an exciting analysis!!! We finally get to answer one of our scientific questions!*

1. Calculated biodiversity measures (Hill numbers) across every sample. 
2. Figures to include in our scientific paper that visualize the data as it relates to the scientific question.
3. Statistical tests conveying the measured and quantified changes and patterns in biodiversity.

# Scientific Question

## Question & Hypotheses

How is microbial diversity and short chain fatty acid concentrations (scfa) modified by inulin treatment.

- *Null Hypothesis:* Microbial diversity and scfa production does not vary during inulin treatment
- *Alternative Hypothesis:* Microbial diversity and scfa production will increase during an inulin treatment. Establishing a correlation between fermentable fiber and gut biota.

# Set up the Environment 

## Timing of this script

```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
```

## Set the seed 
```{r set-seed}
set.seed(3980)
```

## Load Packages & Colors 

```{r load-packages-colors}
# install packages for stats
#install.packages("rstatix")
#install.packages("ggpubr")

pacman::p_load(tidyverse, devtools, patchwork, iNEXT, phyloseq,
               # packages for stats
               ggpubr, rstatix, install = FALSE)

# Load in colors for plotting 
source("code/colors.R")
```

## Load in Data 

```{r load-data}
load("data/04_Phylogenetic_Tree/archaeal_rooted_physeq.RData")

# take a look at it! 
archaeal_rooted_physeq

# Intuition check on seq depth
min(sample_sums(archaeal_rooted_physeq))

# Make a metadata dataframe 
metadata_df <- 
  archaeal_rooted_physeq %>%
  sample_data() %>%
  data.frame()

# view of it
glimpse(metadata_df)
```

## Run iNEXT

```{r iNEXT-biodiversity}
# prepare the input data 
# Species in rows and the samples in columns 
iNEXT_input_df <- 
  archaeal_rooted_physeq %>%
  otu_table() %>%
  data.frame()

# Inpect 
dim(iNEXT_input_df)
iNEXT_input_df[1:5, 1:5]

# Run iNEXT 
# Remember to set the seed!
# ASVs in ROWs, samples in COLUMNS 
#iNEXT_data <- iNEXT(iNEXT_input_df, 
 #                   q = c(0, 1, 2),
 #                   datatype = "abundance")
load("data/05_Biodiversity/iNEXT_data.RData")
# Inspect 
str(iNEXT_data)
typeof(iNEXT_data)

#save(iNEXT_data, file = "data/05_Biodiversity/iNEXT_data.RData")
```

# Rarefaction Curves 

## `ggiNEXT` Rarefaction Curves 

```{r iNEXT-colors}
# Prepare Colors 
color_df <- 
  iNEXT_input_df %>%
  colnames() %>%
  data.frame()

# Check
head(color_df)

# Rename the column 
colnames(color_df)[1] <- "names"
# Check
head(color_df)

# Rename the column 
colnames(color_df)[1] <- "names"
# Check
head(color_df)

# Make a helper dataframe for plotting with colors 
iNEXT_color_df <- 
  color_df %>%
  # Merge with metadata
  left_join(metadata_df, by = "names") %>%
  # Merge with colors for plotting with ggiNEXT
  left_join(data.frame(status_colors = status_colors,
            status = names(status_colors)),
            by = "status")

head(iNEXT_color_df)
```

```{r plot-inext-rarefaction}
ggiNEXT(iNEXT_data, type = 1, facet.var = "Order.q") + 
  scale_color_manual(values = iNEXT_color_df$status_colors) + 
  scale_fill_manual(values = iNEXT_color_df$status_colors) + 
  scale_shape_manual(values = base::rep(17, nsamples(archaeal_rooted_physeq))) + 
  theme_bw() + 
  theme(legend.position = "none") + 
  labs(y = "Effective Number of ASVs", x = "Number of Sequences") 
```
## Manual Rarefaction Curves 

```{r manual-rarefaction}
str(iNEXT_data)

iNEXT_manual_df <- 
  iNEXT_data$iNextEst$size_based %>%
  dplyr::rename(names = Assemblage) %>%
  # join with the metadata
  left_join(., metadata_df, by = "names") %>%
  mutate(status = factor(status, levels = c("before",
                                              "during")))

# Inspection
dim(iNEXT_manual_df)

# Manually plot rarefaction curves 
iNEXT_manual_df %>%
  # Filter out rows that are calculated with actual sequences (no extrapolated)
  dplyr::filter(Method == "Rarefaction") %>%
  # Now, let's draw the plot! 
  ggplot(aes(x = m, y = qD, color = status, group = names)) + 
  geom_line() + 
  # Facet by status 
  facet_grid(Order.q~status, scales = "free") + 
  scale_color_manual(values = status_colors) + 
  theme_bw() + 
  labs(y = "Effective Number of ASVs", x = "Number of Sequences") + 
  theme(legend.position = "none")
```
#Interpretation 1:
Sequencing depth is likely sufficient, all curves plateau even in richness. There is an outlier sample during the treatment that has an unusually high amount of ASVs, I will need to be in the lookout for that.

*Run Shapiro-Wilk for the scf*

```{r shapiro-wilk-scf}

# Test of the data is normal for the continuous value of butyrate
shapiro.test(metadata_df$Butyrate)
# Test of the data is normal for the continuous value of acetate
shapiro.test(metadata_df$acetate)
# Test of the data is normal for the continuous value of propionate
shapiro.test(metadata_df$propionate)
# Test of the data is normal for the continuous value of total_scfa
shapiro.test(metadata_df$total_scfa)
```
Now for the concentration of the 3 scf and total scfa, all of them had p-value<0.05 so none of them are normal, now we can look into richness
```{r shapiro-wilk-richness}
### Now what about for richness? 
obs_div_df <- 
  iNEXT_manual_df %>%
  dplyr::filter(Method == "Observed") 

# check it
glimpse(obs_div_df)

# Pull out unique data from the three fractions of samples 
obs_whole_rich_df <- 
  obs_div_df %>%
  dplyr::filter(Order.q == 0)

# Test of the data is normal for the continuous value of richness
shapiro.test(obs_whole_rich_df$qD)
```
Now richness is not normal, so we need to use non parametric tests.

###I have paired data (multiple data from the same host before and during) so I am going to run a repeated measures Anova, first prunning the NA values for repeatability. Will run chunks per SCFA type and total_scfa

#Total SCFA RM-Anova
```{r total-scfa-RM, fig.height=4, fig.width=4}
# Prune and average total_scfa 
## Prunning breaks the repeated structure of the data, so I will average the different samples per subject into one to approximate the rm ANOVA design.
scfa_summary_df_total_scfa <- metadata_df %>%
  filter(!is.na(total_scfa)) %>%
  group_by(host_subject_id, status) %>%
  summarise(mean_total_scfa = mean(total_scfa), .groups = "drop") %>%
  pivot_wider(names_from = status, values_from = mean_total_scfa) %>%
  drop_na()
# Reshape for Anova
scfa_long_df_total_scfa <- scfa_summary_df_total_scfa %>%
  pivot_longer(cols = c(before, during), names_to = "status", values_to = "total_scfa")
# Repeated measures Anova
scfa_aov_total_scfa <- aov(total_scfa ~ status + Error(host_subject_id), data = scfa_long_df_total_scfa)
summary(scfa_aov_total_scfa)

# Check rigidity of the model with residuals
resid_vals_total_scfa <- residuals(scfa_aov_total_scfa[["Within"]])
#Check normality of residuals
shapiro.test(resid_vals_total_scfa)

# differences in before and after
mean_before_total_scfa <- mean(scfa_summary_df_total_scfa$before)
mean_during_total_scfa <- mean(scfa_summary_df_total_scfa$during)
abs_change_total_scfa <- mean_during_total_scfa - mean_before_total_scfa
sd_before_total_scfa <- sd(scfa_summary_df_total_scfa$before)
sd_during_total_scfa <- sd(scfa_summary_df_total_scfa$during)
percent_change_total_scfa <- 100 * abs_change_total_scfa / mean_before_total_scfa

mean_before_total_scfa
mean_during_total_scfa
sd_before_total_scfa
sd_during_total_scfa
percent_change_total_scfa
```

# acetate RM-Anova
```{r acetate-RM, fig.height=4, fig.width=4}
# Prune and average acetate
## Prunning breaks the repeated structure of the data, so I will average the different samples per subject into one to approximate the rm ANOVA design.
scfa_summary_df_acetate <- metadata_df %>%
  filter(!is.na(acetate)) %>%
  group_by(host_subject_id, status) %>%
  summarise(mean_acetate = mean(acetate), .groups = "drop") %>%
  pivot_wider(names_from = status, values_from = mean_acetate) %>%
  drop_na()
# Reshape for Anova
scfa_long_df_acetate <- scfa_summary_df_acetate %>%
  pivot_longer(cols = c(before, during), names_to = "status", values_to = "acetate")
# Repeated measures Anova
scfa_aov_acetate <- aov(acetate ~ status + Error(host_subject_id), data = scfa_long_df_acetate)
summary(scfa_aov_acetate)

# Check rigidity of the model with residuals
resid_vals_acetate <- residuals(scfa_aov_acetate[["Within"]])
#Check normality of residuals
shapiro.test(resid_vals_acetate)

# differences in before and after
mean_before_acetate <- mean(scfa_summary_df_acetate$before)
mean_during_acetate <- mean(scfa_summary_df_acetate$during)
abs_change_acetate <- mean_during_acetate - mean_before_acetate
sd_before_acetate <- sd(scfa_summary_df_acetate$before)
sd_during_acetate <- sd(scfa_summary_df_acetate$during)
percent_change_acetate <- 100 * abs_change_acetate / mean_before_acetate

mean_before_acetate
mean_during_acetate
sd_before_acetate
sd_during_acetate
percent_change_acetate
```
# Butyrate RM-Anova
```{r Butyrate-RM, fig.height=4, fig.width=4}
# Prune and average Butyrate
## Prunning breaks the repeated structure of the data, so I will average the different samples per subject into one to approximate the rm ANOVA design.
scfa_summary_df_Butyrate <- metadata_df %>%
  filter(!is.na(Butyrate)) %>%
  group_by(host_subject_id, status) %>%
  summarise(mean_Butyrate = mean(Butyrate), .groups = "drop") %>%
  pivot_wider(names_from = status, values_from = mean_Butyrate) %>%
  drop_na()
# Reshape for Anova
scfa_long_df_Butyrate <- scfa_summary_df_Butyrate %>%
  pivot_longer(cols = c(before, during), names_to = "status", values_to = "Butyrate")
# Repeated measures Anova
scfa_aov_Butyrate <- aov(Butyrate ~ status + Error(host_subject_id), data = scfa_long_df_Butyrate)
summary(scfa_aov_Butyrate)

# Check rigidity of the model with residuals
resid_vals_Butyrate <- residuals(scfa_aov_Butyrate[["Within"]])
#Check normality of residuals
shapiro.test(resid_vals_Butyrate)

# differences in before and after
mean_before_Butyrate <- mean(scfa_summary_df_Butyrate$before)
mean_during_Butyrate <- mean(scfa_summary_df_Butyrate$during)
abs_change_Butyrate <- mean_during_Butyrate - mean_before_Butyrate
sd_before_Butyrate <- sd(scfa_summary_df_Butyrate$before)
sd_during_Butyrate <- sd(scfa_summary_df_Butyrate$during)
percent_change_Butyrate <- 100 * abs_change_Butyrate / mean_before_Butyrate

mean_before_Butyrate
mean_during_Butyrate
sd_before_Butyrate
sd_during_Butyrate
percent_change_Butyrate
```
#Propionate RM-Anova
```{r Propionate-RM, fig.height=4, fig.width=4}
# Prune and average propionate
## Prunning breaks the repeated structure of the data, so I will average the different samples per subject into one to approximate the rm ANOVA design.
scfa_summary_df_propionate <- metadata_df %>%
  filter(!is.na(propionate)) %>%
  group_by(host_subject_id, status) %>%
  summarise(mean_propionate = mean(propionate), .groups = "drop") %>%
  pivot_wider(names_from = status, values_from = mean_propionate) %>%
  drop_na()
# Reshape for Anova
scfa_long_df_propionate <- scfa_summary_df_propionate %>%
  pivot_longer(cols = c(before, during), names_to = "status", values_to = "propionate")
# Repeated measures Anova
scfa_aov_propionate <- aov(propionate ~ status + Error(host_subject_id), data = scfa_long_df_propionate)
summary(scfa_aov_propionate)

# Check rigidity of the model with residuals
resid_vals_propionate <- residuals(scfa_aov_propionate[["Within"]])
#Check normality of residuals
shapiro.test(resid_vals_propionate)

# differences in before and after
mean_before_propionate <- mean(scfa_summary_df_propionate$before)
mean_during_propionate <- mean(scfa_summary_df_propionate$during)
abs_change_propionate <- mean_during_propionate - mean_before_propionate
sd_before_propionate <- sd(scfa_summary_df_propionate$before)
sd_during_propionate <- sd(scfa_summary_df_propionate$during)
percent_change_propionate <- 100 * abs_change_propionate / mean_before_propionate

mean_before_propionate
mean_during_propionate
sd_before_propionate
sd_during_propionate
percent_change_propionate
```
#Plotting the total scfa
```{r-plot-total-scfa}
# Salinity by status 
total_scfa_ggboxplot <- 
  ggboxplot(scfa_long_df_total_scfa, x = "status", y = "total_scfa", 
          color = "status", fill = "status", alpha = 0.3,
          outlier.shape = NA) + 
  geom_jitter(aes(color = status)) + 
  scale_color_manual(values = status_colors) + 
  scale_fill_manual(values = status_colors) + 
  labs(y = "total SCFA (mmol/kg)", subtitle = "Repeated-Measures ANOVA: F(1,21) = 0.89, p = 0.356") +
  theme_bw() + 
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1))
#Show the plot
total_scfa_ggboxplot

```
The graph shows that there are no significant differences between the boxplots, the median before is slightly higher.

### Diversity by status
```{r-make-obs-div}
# Plot boxplots by status against diversity
obs_div_df %>%
  ggplot(aes(x = status, y = qD, fill = status, 
             color = status)) + 
  facet_wrap(.~Order.q, scales = "free_y") + 
  geom_jitter() + 
  geom_boxplot(alpha = 0.5, outlier.shape = NA) + 
  scale_color_manual(values = status_colors) + 
  scale_fill_manual(values = status_colors) + 
  labs(y = "Effective Number of ASVs") + 
  theme_bw() + 
  theme(legend.position = "bottom",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 30,
                                   hjust = 1,
                                   vjust = 1))
```


#### Statistically test Richness
```{r stat-richness}
# Pull out richness data 
obs_rich_df <- 
  obs_div_df %>%
  dplyr::filter(Order.q == 0)

kw_status_rich <- 
  obs_rich_df %>%
  kruskal_test(qD ~ status)

# View results
kw_status_rich
summary(kw_status_rich)

rich_ggboxplot <-
  ggboxplot(obs_rich_df, x = "status", y = "qD", fill = "status", 
             color = "status", alpha = 0.3,
          outlier.shape = NA) + 
 geom_jitter(aes(color = status)) + 
  scale_color_manual(values = status_colors) + 
  scale_fill_manual(values = status_colors) + 
  labs(y = "# of Total ASVs (Richness)") + 
  stat_compare_means(method = "kruskal.test", label.y = min(obs_rich_df$qD)) + 
  geom_pwc(
    aes(group = status), tip.length = 0, hide.ns = TRUE,
    method = "t_test", label = "p.adj.format",
    p.adjust.method = "fdr", bracket.nudge.y = -0.08) + 
  theme_bw() + 
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 30,
                                   hjust = 1, 
                                   vjust = 1))
# Pull out simpson data 
obs_simpson_df <- 
  obs_div_df %>%
  dplyr::filter(Order.q == 2)
# Kruskal walis test
kw_simpson_df <- 
  obs_rich_df %>%
  kruskal_test(qD ~ status)

# View results
kw_simpson_df
summary(kw_simpson_df)   

simpson_ggboxplot <-
  ggboxplot(obs_simpson_df, x = "status", y = "qD", fill = "status", 
             color = "status", alpha = 0.3,
          outlier.shape = NA) + 
 geom_jitter(aes(color = status)) + 
  scale_color_manual(values = status_colors) + 
  scale_fill_manual(values = status_colors) + 
  labs(y = "# of dominant ASVs (simpson)") + 
  stat_compare_means(method = "kruskal.test", label.y = 50) + 
  geom_pwc(
    aes(group = status), tip.length = 0, hide.ns = TRUE,
    method = "t_test", label = "p.adj.format",
    p.adjust.method = "fdr", bracket.nudge.y = -0.08) + 
  theme_bw() + 
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 30,
                                   hjust = 1, 
                                   vjust = 1))

# Show the plots
rich_ggboxplot + simpson_ggboxplot
```
#Interpretation 2:
I can conclude based on the kruskal-wallis tests that rare taxa change based on the inulin treatment but dominant taxa stay relatively stable with a non-significant change.
#Interpretation 3: 
The kruskall-wallis test is testing for a difference in microbial composition before and after the inulin treatment, a significant value in richness means that rare taxa are affected by a dietary intervention with fermentable fibers, although dominant species maintain stable.

# scfa concentrations: 
```{r-scfa-concentrations}
# Create dataframe of scfa concentrations
scfa_df <- 
  metadata_df %>%
  dplyr::select(Experiment, acetate, Butyrate, propionate, total_scfa)

# Pair-wise interactions between scfas
pairs(dplyr::select(scfa_df, -Experiment), upper.panel = NULL)
```

#How does total_scfa change in relation do diversity
```{r plot-biodiversity-scfa}
# How does biodiversity change over the salinity gradient?
obs_div_df %>%
  ggplot(aes(x = total_scfa, y = qD)) + 
  facet_wrap(.~Order.q, scales = "free") + 
  geom_point(aes(color = status)) + 
  scale_color_manual(values = status_colors) + 
  labs(x = "total_scfa (millimoles per kilogram)", y = "Effective Number of ASVs") +
  stat_smooth(method = "lm", formula = y ~poly(x, 2)) + 
  theme_bw() + 
  theme(legend.position = "bottom",
        legend.title = element_blank())
```
There appears to be a negative relationship between scfa concentration and effective number of ASVs in the common and dominant, while the richness stays stable ( mainly in the extrapolations the actual data does not reflect clear trends)

#Explore more richness
```{r richness-stat}
# Make a richness data frame
richness_df <- 
  obs_div_df %>%
  dplyr::filter(Method == "Observed") %>%
  dplyr::filter(Order.q == 0)

# Actually run the linear model
rich_vs_total_scfa <- 
  lm(qD ~ total_scfa, data = richness_df)

# Show me the model results
rich_vs_total_scfa
# Type?
class(rich_vs_total_scfa)
# Summary of lm
summary(rich_vs_total_scfa)
```

#Richness plot

```{r richness-plot}
richness_df %>% 
  ggplot(aes(x = total_scfa, y = qD)) + 
  geom_point(aes(color = status)) + 
  stat_smooth(method = "lm") + 
  labs(x = "Salinity (PSU)", y = "# of ASVs") + 
  scale_color_manual(values = status_colors) + 
  theme_bw() + 
  theme(legend.position = "bottom",
        legend.title = element_blank()) + 
  geom_label(aes(x = 20, y = 60), hjust = 0, 
             label = paste("Adj R2 = ",signif(summary(rich_vs_total_scfa)$adj.r.squared, 2),
                           "\nIntercept =",signif(rich_vs_total_scfa$coef[[1]],3),
                           " \nSlope =",signif(rich_vs_total_scfa$coef[[2]], 2),
                           " \nP =",signif(summary(rich_vs_total_scfa)$coef[2,4], 2)))

```

#Spearman correlation

```{r-spearman-correl}

richness_df %>% 
  ggplot(aes(x = total_scfa, y = qD)) + 
  geom_point(aes(color = status)) + 
  stat_smooth(method = "lm") + 
  labs(x = "total SCFA", y = "# of ASVs") + 
  scale_color_manual(values = status_colors) + 
  theme_bw() + 
  theme(legend.position = "bottom",
        legend.title = element_blank()) + 
  stat_cor(method = "spearman", label.x = 3, label.y = 30)
```
#Interpretation 4: 
Richness and rare taxa were the only ones that changed statistically significantly but it looks like total_scfa is not able to significantly explain this change, so scfa production does not relate to changes in richness significantly.

#Interpretation 5:
I checked it by asking if total_scfa could explain the changes in richness observed during the kruskall wallis test, because a part of my scientific question was to determine whether scfa concentrations changed with diversity after the inulin treatment.

# Final info for Reproducibility 

## Check Render Time
```{r stop-time}
# Take the time now that we are at the end of the script
end_time <- Sys.time()
end_time 

# Echo the elapsed time
elapsed_time <- round((end_time - start_time), 3)
elapsed_time
```

## Session Information

```{r session-info}
# Ensure reproducibility with package version information
devtools::session_info()
```
