---
output: html_document
editor_options: 
  chunk_output_type: console
---
title: "Project ASVs assignment"
author: "Roger Calvas"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      fig.path = "figures/01_DADA2")
```
# Set the seed 
```{r set-seed}
# any number can be chosen
set.seed(3980)
```
### Timing of Script
```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
start_time
```
#Goals of this file:

1. Load in the raw sequencing data.
2. Visualize and inspect sequencing data to asess the quality of the reads.
3. Filter by the quality and trim primers from sequencing data.
4. Write out new fastq files that include high quality and trimmed sequences.
5. Re-inspect the quality of the reads. (intuition check)
6. Infer the error rate of the forward (R1) and reverse (R2) reads.
**DADA2 building the error model**
7. Apply the error model and infer ASVs on forward and reverse reads individually.
8. Merge forward and reverse ASVs into "contiguous ASVs".
9. Generate an ASV count table (`otu_table` for phyloseq.)

output that we need: 

1. ASV count table: `otu_table`
2.Quality plots to asses pre- and post-quality trimming.
3. Written files that include the filtered and trimmed data

#load R libraries
```{r}
#Efficiently load with pacman
pacman::p_load(tidyverse, dada2, phyloseq, DT, devtools, install = FALSE)
library("patchwork")
```

# Load in the raw sequencing data.
```{r}
#set the path 
raw_fastqs_path <- "data/fastq_data/fall15"

#what files are in the path?
head(list.files(raw_fastqs_path))
#how many are there 
length(list.files(raw_fastqs_path))

#create a vector of forward reads (R1)
forward_reads <- list.files(raw_fastqs_path, pattern = "1.fastq.gz", full.names = TRUE)

#intuition check
stopifnot(length(forward_reads) < length(list.files(raw_fastqs_path)))

#create a vector of reverse reads (R2)
 reverse_reads <-list.files(raw_fastqs_path, pattern = "2.fastq.gz", full.names = TRUE)
 #intuition check
stopifnot(length(reverse_reads) < length(list.files(raw_fastqs_path)))
stopifnot(length(reverse_reads) == length(forward_reads))
```

# Visualize and inspect sequencing data to asess the quality of the reads.
```{r raw-quality-plot, fig.width=12, fig.height=8}
# Randomly select 12 samples 
random_samples <- sample(1:length(reverse_reads), size = 12)
random_samples

#Calculate and plot the quality of these 12 samples

#forward quality plot
forward_filteredQual_plot_12 <- plotQualityProfile(forward_reads[random_samples]) + 
  labs(title = "Forward Read: Raw Quality")

#reverse quality plot
reverse_filteredQual_plot_12 <- plotQualityProfile(reverse_reads[random_samples]) + 
  labs(title = "Reverse Read: Raw Quality")

#put the two blocks together
forward_filteredQual_plot_12 + reverse_filteredQual_plot_12

```

[The quality seems good enough, overall there is a big difference in quality between the forward and reverse reads after 170~ with the difference being more noticeable at 200 bases, because this dips in quality seem to be randomly spread across the reads truncation may need to occur per phred score as to not loss almost half of our length]


#aggregate al qc plots

```{r aggregate all qc plots }

# Aggregate all QC plots 
# Forward reads
forward_preQC_plot <- 
  plotQualityProfile(forward_reads, aggregate = TRUE) + 
  labs(title = "Forward Pre-QC")

# reverse reads
reverse_preQC_plot <- 
  plotQualityProfile(reverse_reads, aggregate = TRUE) + 
  labs(title = "Reverse Pre-QC")

# Now, let's put the two plots together
preQC_aggregate_plot <- 
  # Plot the forward and reverse together 
  forward_preQC_plot + reverse_preQC_plot
# Show the plot
preQC_aggregate_plot
```
[Early Bases have a drop in quality (~Cycle/Bases 0-10):
Both forward and reverse reads show a steep quality drop in the first ~10 bases, but this is mostly above 30, which is acceptable, so no trimming is needed.
High Quality in Mid-Read for forward(1-150~ Cycles/Bases):
After 150 bases the quality starts to decline mainly above 30 before the 220~ base.
Further decline in Forward Reads After ~220 Cycles/Bases:
After 220~ bases the quality starts to have some dips going below 30 up to 25 qc points, where some of the data is usable, since is such a big part of the sequence I am deciding to keep it.
Acceptable Quality in Mid-Read for reverse(1-170~ Cycles/Bases):
Between 1 and 170~ bases there are a lot of dips but most staying above 30, after 170~ bases a steep decline in quality starts.
Further decline in reverse Reads After ~170 Cycles/Bases:
After 170~ the quality of the reads start to diminish with some dips below 30, after 200~ bases this gets amplified and there are a lot of dips below 20 making a lot of the data after this unusable.
The last couple of bases in the reverse read present the typical dip into unusable quality.
This effect was seen in fastqc and multiqc, with a major dip in quality by the end of the run.
Solution: We will use stricter expected error rates for the reverse read as to maintain as most depth as possible for future analysis, then truncLen = c(250,200) to eliminate the sudden abrupt dip in quality for the reverse length.

Forward Reads Maintain Quality longer and in average have a higher fidelity overall.
Solution: The reads will be exposed to different maxEE values as to eliminate the least depth as possible
Forward reads degrade after ~170 cycles.
Reverse reads have a worst quality overall requiring stricter controls for appropriate quality.]
# Filter by the quality and trim primers from sequencing data.

####  Prepare a placeholder for filtered reads 

```{r prep-filtered-sequences}
# Create vector of sample names from the filenames 
sample_names <- sapply(strsplit(basename(forward_reads), "_"), `[`,1) 

# Intuition Check 
head(sample_names)

# Place filtered reads into filtered_fastqs_path
filtered_fastqs_path <- "data/01_DADA2/02_filtered_fastqs"

# Intuition Check 
filtered_fastqs_path

# create 2 vectors: filtered_forward_reads & filtered_reverse_reads
filtered_forward_reads <- 
  file.path(filtered_fastqs_path, paste0(sample_names, "_R1_filtered.fastq.gz"))

# Intuition Check 
length(filtered_forward_reads)

# reverse reads
filtered_reverse_reads <- 
  file.path(filtered_fastqs_path, paste0(sample_names, "_R2_filtered.fastq.gz"))

# Intuition Check 
head(filtered_reverse_reads)
```
## Filter and Trim Reads
```{r filter-and-trim}
filtered_reads <- 
  filterAndTrim(fwd = forward_reads, filt = filtered_forward_reads,
              rev = reverse_reads, filt.rev = filtered_reverse_reads,
              maxN = 0, maxEE = c(1,1), 
              truncLen = c(240,217),
              trimLeft = 4,
              truncQ = 2,
              rm.phix = TRUE, compress = TRUE,
              multithread = 10)

```

# Plot the 12 random samples after QC

```{r-plotting the trimmed sequences}

forward_filteredQual_plot_12 <- 
  plotQualityProfile(filtered_forward_reads[random_samples]) + 
  labs(title = "Trimmed Forward Read Quality")

reverse_filteredQual_plot_12 <- 
  plotQualityProfile(filtered_reverse_reads[random_samples]) + 
  labs(title = "Trimmed Reverse Read Quality")

# Put the two plots together 
forward_filteredQual_plot_12 + reverse_filteredQual_plot_12
```

# Aggregate all QC plots

```{r Agregate-postqc plot}
# Forward reads
forward_postQC_plot <- 
  plotQualityProfile(filtered_forward_reads, aggregate = TRUE) + 
  labs(title = "Forward Post-QC")

# reverse reads
reverse_postQC_plot <- 
  plotQualityProfile(filtered_reverse_reads, aggregate = TRUE) + 
  labs(title = "Reverse Post-QC")

# Now, let's put the two plots together
postQC_aggregate_plot <- 
  # Plot the forward and reverse together 
  forward_postQC_plot + reverse_postQC_plot
# Show the plot
postQC_aggregate_plot
```
[The random dips in quality have been mostly eliminated while maintaining good sequencing depth, the dip at the end of the reverse read has been eliminated as well.
The quality of the sequences has become a lot better with no sequences dipping below 26~]
### Visualize QC differences in plot 
```{r pre-post-QC-plot, fig.width=6, fig.height=5.5}
# Plot the pre and post together in one plot
preQC_aggregate_plot / postQC_aggregate_plot
```
# Make output into dataframe

```{r data-frame}
 
filtered_df <- as.data.frame(filtered_reads) %>%
  mutate(percent.retained = reads.out/reads.in)

# Intuition check
# Visualize it in table format 
DT::datatable(filtered_df)
```

#calculate statistics
```{r statistics}
read_stats_df <- 
  filtered_df %>%
  reframe(median_reads_in = median(reads.in),
          median_reads_out = median(reads.out),
          median_percent_retained = (median(reads.out)/median(reads.in)),
          max_percent_retained = max(reads.out/reads.in),
          min_percent_retained = min(reads.out/reads.in))

# Take a look at it!
read_stats_df
```
plotting stats
```{r plot-stats}
numSeqs_QC_dotplot <-
  filtered_df %>%
  ggplot(aes(x = reads.in, y = reads.out)) + 
  geom_point(alpha = 0.5, size = 2) + 
  labs(x = "# of Raw Seqs", 
       y = "# of Seqs Retained") + 
  # Now let's add a 1:1 line for reference of keeping 100% of the reads
  geom_abline(slope=1, intercept = 0, color = "deeppink")

# Now, let's look at the number of reads retained in a histogram
numRetained_QC_histplot <- 
  filtered_df %>%
  ggplot(aes(x = reads.out)) + 
  geom_histogram() + 
  labs(x = "# of Seqs Retained", 
       y = "# of Samples") 

# Create a histogram of percent reads retained in a histogram
percSeqs_QC_histplot <- 
  filtered_df %>%
  ggplot(aes(x = percent.retained)) + 
  geom_histogram() + 
  labs(x = "% of Seqs Retained", 
       y = "# of Samples") + 
  # Set the scale to be between 0-1 (0-100%)
  scale_x_continuous(limits = c(0, 1))

# Now, let's put the plots together
numSeqs_QC_dotplot + numRetained_QC_histplot + percSeqs_QC_histplot + 
  plot_annotation(tag_levels = 'A')

```
[Most of the sequences were maintained and general distribution of the seqs across samples was also mantained]
